---
title: "Lab 34: Scientific Poster Preparation Workshop"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "November 23, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Communicating Science Visually

A scientific poster is not a printed version of a paper. It is a **visual conversation starter** — designed to be understood in 3 minutes by someone standing a metre away, and to invite the reader to ask you questions. This lab covers the principles of poster design, how to produce publication-quality figures in R, and how to use a peer review process to improve your work before the final presentation.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Identify the key sections of a scientific poster and what goes in each
2. Apply design principles: visual hierarchy, colour, whitespace, and font size
3. Produce publication-quality R figures at 300 dpi with appropriate dimensions
4. Write a poster title, summary, and conclusion that communicate to a non-specialist
5. Give and receive constructive peer feedback using a structured rubric
6. Export figures from R in formats suitable for PowerPoint and printing (PNG, PDF, SVG)
:::

------------------------------------------------------------------------

## Part 1: Anatomy of a Scientific Poster

::: callout-important
## Standard Poster Sections

**Title bar** — spans the full width. Contains: title (largest text on poster), your name, course/institution, date.

**Introduction / Background** — 3–5 sentences max. What question are you answering and why does it matter? No citations needed at this level.

**Methods** — brief description of your data and analysis approach. Can use a numbered list or a simple flowchart.

**Results** — this is the heart of the poster. 2–4 figures, each self-contained with a caption. The figures should tell the story without extensive text.

**Discussion / Conclusion** — 3–5 bullet points or 2 short paragraphs. What do your results mean? Any limitations?

**References** — 3–5 key references if appropriate.

**Acknowledgements** — brief (1–2 lines).
:::

```{r, echo=FALSE}
cat('
POSTER LAYOUT GUIDE (A0 landscape = 1189 x 841 mm)
====================================================

[          TITLE / NAME / INSTITUTION               ]
[                                                    ]
[ Intro  ] [   Results Fig 1  ] [  Results Fig 3   ]
[        ] [                  ] [                  ]
[ Methods] [   Results Fig 2  ] [  Discussion      ]
[        ] [                  ] [                  ]
[        ] [   (main figure)  ] [  Conclusion      ]

DESIGN RULES:
  - Title: 80–100pt font
  - Body text: minimum 24pt (readable from 1 metre)
  - Maximum 500 words of text on entire poster
  - 30–40% whitespace (empty space is not wasted space)
  - Maximum 4 colours (including black/white)
  - Every figure must have a self-contained caption
')
```

------------------------------------------------------------------------

## Part 2: Producing Publication-Quality Figures

The key difference between exploratory figures (for your own analysis) and publication/poster figures is:

- Font size large enough to read when printed
- Clean theme (no grey background)
- Appropriate dimensions for the intended space
- Saved at 300 dpi (print quality) or as a vector format (PDF/SVG)

### Example: Polishing a Figure for a Poster

```{r}
library(ggplot2)
library(dplyr)

# Use a dataset that might appear in a BI 255 poster
# Enzyme activity across three treatment groups
set.seed(42)
poster_df <- data.frame(
  treatment = rep(c("Control", "Low Dose", "High Dose"), each = 20),
  activity  = c(rnorm(20, 50, 8), rnorm(20, 72, 10), rnorm(20, 45, 12))
)
poster_df$treatment <- factor(poster_df$treatment,
                               levels = c("Control", "Low Dose", "High Dose"))

# Version 1: Default ggplot (NOT suitable for poster)
ggplot(poster_df, aes(x = treatment, y = activity, fill = treatment)) +
  geom_boxplot() +
  labs(title = "Enzyme Activity by Treatment")
```

```{r}
# Version 2: Publication-quality poster figure
poster_fig <- ggplot(poster_df, aes(x = treatment, y = activity, fill = treatment)) +
  geom_violin(trim = FALSE, alpha = 0.55) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA, linewidth = 0.8) +
  geom_jitter(width = 0.08, size = 2, alpha = 0.6, colour = "grey30") +
  scale_fill_manual(values = c("Control"   = "#A8C4D9",
                               "Low Dose"  = "#2C5F8A",
                               "High Dose" = "#C8102E")) +
  labs(
    title   = "Enzyme Activity Is Significantly Altered by Treatment",
    x       = NULL,
    y       = "Enzyme Activity (nmol/min/mg protein)",
    caption = "n = 20 per group; one-way ANOVA, Tukey post-hoc (not shown)"
  ) +
  theme_classic(base_size = 18) +   # Large base font for poster
  theme(
    legend.position  = "none",
    plot.title       = element_text(size = 16, face = "bold"),
    axis.text        = element_text(size = 14),
    axis.title.y     = element_text(size = 14),
    plot.caption     = element_text(size = 10, colour = "grey50")
  )

poster_fig
```

```{r, eval=FALSE}
# Save for printing (300 dpi PNG) and for importing into PowerPoint
ggsave("enzyme_poster_figure.png",
       plot   = poster_fig,
       width  = 18,    # cm — size in the poster layout
       height = 14,
       units  = "cm",
       dpi    = 300)

# Vector format (PDF or SVG — infinitely scalable, ideal for posters)
ggsave("enzyme_poster_figure.pdf",
       plot   = poster_fig,
       width  = 18, height = 14, units = "cm")
```

::: callout-tip
## Figure Size and Resolution

| Format | When to use |
|--------|-------------|
| PNG at 300 dpi | Standard for posters and papers; fixed resolution |
| PDF | Vector — perfectly sharp at any size; for InDesign or Illustrator |
| SVG | Vector — works in PowerPoint and web; editable in Inkscape |
| TIFF at 300+ dpi | Some journals require this for submission |

For a PowerPoint poster: use PNG at 300 dpi. For a professionally printed poster: use PDF or high-resolution PNG (300 dpi minimum; 600 dpi preferred for text-heavy figures).
:::

------------------------------------------------------------------------

## Part 3: Writing Poster Text

```{r, echo=FALSE}
cat('
EXAMPLE: TRANSFORMING RESULTS TEXT FOR A POSTER
================================================

PAPER VERSION (too dense for a poster):
  "Enzyme activity was significantly elevated in the low-dose group
   (72.3 ± 10.1 nmol/min/mg) compared to both the control group
   (50.4 ± 8.3 nmol/min/mg; p = 0.003) and the high-dose group
   (45.2 ± 11.8 nmol/min/mg; p < 0.001), as determined by one-way
   ANOVA (F(2,57) = 18.4, p < 0.001) followed by Tukey-Kramer
   post-hoc comparisons."

POSTER VERSION (concise and visual):
  Title of figure: "Low-Dose Treatment Maximises Enzyme Activity"
  Caption: "n = 20 per group. One-way ANOVA: F(2,57) = 18.4, p < 0.001.
            Letters indicate Tukey-Kramer post-hoc groupings."
  The figure shows the result. The text reinforces it in one sentence.

KEY RULE: If the figure caption already explains the result,
the body text does not need to repeat it. Point to the figure.
')
```

### Titles That Tell the Story

```{r, echo=FALSE}
cat('
WEAK TITLE (describes what you did):
  "Enzyme Activity Results Under Three Concentrations"

STRONG TITLE (tells the main finding):
  "Low-Dose Treatment Maximises Enzyme Activity — High Dose Is Inhibitory"

A strong poster title is a headline — it gives away the answer.
The reader should know your main finding before reading anything else.
')
```

------------------------------------------------------------------------

## Part 4: Creating a Figure Panel

Many posters include a multi-panel figure. Here is how to produce a clean panel layout in R:

```{r}
# install.packages("patchwork")
library(patchwork)

# Panel A: histogram of control data
fig_A <- ggplot(poster_df |> filter(treatment == "Control"),
                aes(x = activity)) +
  geom_histogram(fill = "#A8C4D9", colour = "white", bins = 10) +
  labs(title = "A. Control Distribution",
       x = "Enzyme Activity", y = "Count") +
  theme_classic(base_size = 14)

# Panel B: scatter-style comparison
fig_B <- ggplot(poster_df, aes(x = treatment, y = activity, colour = treatment)) +
  geom_jitter(width = 0.15, size = 2.5, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.3, colour = "black") +
  scale_colour_manual(values = c("Control" = "#A8C4D9",
                                 "Low Dose" = "#2C5F8A",
                                 "High Dose" = "#C8102E")) +
  labs(title = "B. Group Comparison",
       x = NULL, y = "Enzyme Activity") +
  theme_classic(base_size = 14) +
  theme(legend.position = "none")

# Combine with patchwork
combined_panel <- fig_A | fig_B
combined_panel + plot_annotation(
  title = "Figure 1. Enzyme Activity Under Three Treatment Conditions",
  theme = theme(plot.title = element_text(size = 13, face = "bold"))
)
```

```{r, eval=FALSE}
# Save the panel at poster quality
ggsave("figure1_panel.png",
       combined_panel,
       width = 28, height = 14, units = "cm", dpi = 300)
```

------------------------------------------------------------------------

## Part 5: Peer Review Rubric

Use this rubric to give structured, constructive feedback on a classmate's poster draft.

```{r, echo=FALSE}
cat('
BI 255 POSTER PEER REVIEW RUBRIC
==================================

VISUAL DESIGN (20 pts)
  [ ] Title is readable from 1 metre and states the main finding
  [ ] Font size is minimum 24pt for body text
  [ ] Figures are clearly labelled with axes and units
  [ ] Colour palette is consistent and accessible
  [ ] Whitespace is used well (not overcrowded)

SCIENTIFIC CONTENT (40 pts)
  [ ] Research question is clearly stated in Introduction
  [ ] Methods are described concisely (enough to replicate the analysis)
  [ ] Results section contains 2-4 clear, well-captioned figures
  [ ] Statistical tests are named and results reported (F, t, p, or CI)
  [ ] Conclusions are supported by the data shown

COMMUNICATION (20 pts)
  [ ] Poster tells a clear story from top to bottom
  [ ] Non-specialist could understand the main finding
  [ ] Technical jargon is defined or avoided
  [ ] Conclusions are appropriately qualified (limitations noted)

R FIGURE QUALITY (20 pts)
  [ ] Figures were produced in R (ggplot2 or equivalent)
  [ ] Figures show raw data (not just means/bars)
  [ ] Font size in figures is large enough to read when printed
  [ ] Colour scheme is consistent with poster colour palette

TOTAL: ___ / 100

ONE STRENGTH: _________________________________

ONE SUGGESTION: ________________________________
')
```

------------------------------------------------------------------------

## Part 6: Presentation Tips

::: callout-note
## The 2-Minute Elevator Pitch

When someone stops at your poster, you have approximately 2 minutes before they decide to engage or move on. Practice this structure:

1. **The question** (15 seconds): "My project asked whether..."
2. **The approach** (20 seconds): "I used data from... and analysed it with..."
3. **The main finding** (30 seconds): "The key result was... [point to the main figure]"
4. **Why it matters** (15 seconds): "This is important because..."
5. **Invite a question** (40 seconds): "I'd love to hear your thoughts — especially about..."

Practice this out loud at least five times before the presentation.
:::

```{r, echo=FALSE}
cat('
COMMON MISTAKES TO AVOID
=========================

1. Too much text — aim for a maximum of 500 words total
2. Too many figures — 2 to 4 clear figures beat 8 cluttered ones
3. Bar charts without raw data — show the distribution (violin, jitter)
4. Missing units on axes — always label with units in parentheses
5. Printing at too low resolution — minimum 300 dpi for poster printing
6. Reading the poster out loud during presentation — talk to the person, 
   not the poster
7. Unexplained statistical notation — define p-values, CIs, and test names
8. Inconsistent fonts and colours — pick a palette and stick to it
')
```

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** What is the minimum font size recommended for poster body text, and why?

**2.** You have a beautiful regression plot saved at 72 dpi. The printed poster is A0 size. What problem will occur, and how do you fix it?

**3.** Which ggplot2 theme argument produces figures with the clean appearance most appropriate for a scientific poster?

    a) `theme_grey()`    b) `theme_classic()`    c) `theme_dark()`    d) `theme_void()`

**4.** A strong poster figure caption should include which three elements?

**5.** True or False: Saving a figure as PDF from R means it will look blurry if scaled up on a large poster.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** The minimum is **24pt** for body text on an A0 poster. The rule of thumb is that text should be readable from approximately 1 metre away. Titles should be 80–100pt, section headings 40–48pt, body text 24–28pt, and figure labels/captions at least 18–20pt. Text smaller than 24pt requires the reader to step forward, disrupting the natural flow of the poster session.

**2.** A 72 dpi image will appear **pixelated / blurry** when printed at poster size. At 72 dpi, each pixel is large and visible at print resolution. Fix it by regenerating the figure with `ggsave(..., dpi = 300)` and re-exporting. Alternatively, save as PDF (vector) which is infinitely scalable with no quality loss. Never submit 72 dpi (screen resolution) images for print.

**3.** b) `theme_classic()` — removes the grey background and grid lines, producing a clean white background with axis lines only. This is the standard for published biology figures. `theme_grey()` (the default) has a grey background inappropriate for print; `theme_dark()` is unsuitable; `theme_void()` removes all axes which is only appropriate for maps.

**4.** A strong figure caption should include: (1) **A title sentence** stating the main finding (not just what the figure shows), (2) **Sample size and key experimental details** (n = X per group, conditions, etc.), (3) **Statistical results** (test name, test statistic, degrees of freedom, and p-value) — enough for the reader to judge the conclusion. Example: "Enzyme activity was significantly higher in the low-dose group. n = 20 per group; one-way ANOVA: F(2,57) = 18.4, p < 0.001; Tukey-Kramer post-hoc results indicated by letters."

**5.** False — PDFs saved from R are **vector graphics** — they are mathematically defined curves and shapes with no pixel resolution. Vector images are perfectly sharp at any size, from a business card to a billboard. This is why PDF (or SVG) is the ideal format for poster figures. Raster formats (PNG, JPEG, TIFF) are resolution-dependent and must be saved at 300 dpi or higher for print.
:::

------------------------------------------------------------------------

## Lab 34 Checklist

Before you leave, make sure you can:

- [ ] List the required sections of a scientific poster and what each contains
- [ ] Apply `theme_classic(base_size = 18)` and appropriate colour palettes for poster figures
- [ ] Save a figure at 300 dpi PNG and as PDF using `ggsave()`
- [ ] Create a multi-panel figure with the `patchwork` package
- [ ] Write a poster title that states the main finding (not just the topic)
- [ ] Give structured peer feedback using the rubric
- [ ] Deliver a 2-minute verbal summary of your research

::: callout-tip
## Final Poster Checklist (Take This to Your Printing Appointment)

Before submitting your poster file for printing:

- [ ] All text is at least 24pt
- [ ] All figures are 300 dpi or vector (PDF/SVG)
- [ ] Axes have units in parentheses
- [ ] Statistical results are reported in standard format
- [ ] No more than 4 colours in the whole poster
- [ ] The title states your main finding
- [ ] You have a 2-minute pitch practiced and timed
- [ ] Poster dimensions match the required size (check Canvas)
:::

------------------------------------------------------------------------

## Before the Final Presentation

- Review all figure code from Labs 12–13 (ggplot2), 19 (ANOVA), 22 (regression), and whichever labs relate to your project
- Ensure your Posit Cloud project is saved and backed up
- Prepare questions you can ask the audience to engage them
- Check Canvas for the final poster submission deadline and presentation schedule

------------------------------------------------------------------------
