---
title: "Lab 2: Basic Building Blocks in R"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "August 19, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Learning R with Taylor Swift

Today we're going to learn the fundamental building blocks of R — variables, data types, and basic functions — using data from one of the best-selling music artists of all time: **Taylor Swift** (probably not your favorite, or mine; but a good source of data).

Her albums have sold hundreds of millions of copies, broken streaming records, and generated billions of dollars. That data is perfect for learning R.

::: callout-note
## Learning Objectives

By the end of this lab you will be able to:

1.  Assign values to variables using `<-`
2.  Identify the three main data types: numeric, character, and logical
3.  Use built-in R functions like `print()`, `class()`, `nchar()`, and `paste()`
4.  Perform arithmetic and comparisons on variables
:::

------------------------------------------------------------------------

## Part 1: Variables — Storing Information

In R, a **variable** is a named container that holds a value. You create one using the **assignment operator** `<-`.

Think of it like giving a nickname to a piece of information so you can use it again later.

```{r}
# Taylor Swift's album "1989" sold approximately 10 million copies in its first year
albums_sold <- 10000000

# Print the value
albums_sold
```

```{r}
# Her Eras Tour gross revenue (in USD) as of 2024
eras_tour_revenue <- 2077000000

# You can do math with variables
eras_tour_revenue / 1000000  # Convert to millions
```

::: callout-tip
## ️ Naming Rules for Variables

- Names can contain letters, numbers, `.` and `_`
- Names **cannot** start with a number
- R is **case-sensitive**: `MyVar` and `myvar` are different things
- Use descriptive names: `album_sales` is better than `x`

Good: `album_sales`, `tour_revenue_2024`, `track.count` Bad: `1album`, `my variable`, `x`
:::

------------------------------------------------------------------------

## Part 2: Data Types — What Kind of Information?

Not all information is the same kind. R recognizes different **data types**. The three most important for you right now are:

| Type | What it stores | Example |
|------------------|---------------------------------|---------------------|
| **numeric** | Numbers (with or without decimals) | `3.14`, `42`, `1000000` |
| **character** | Text (always in quotes) | `"Taylor Swift"`, `"Folklore"` |
| **logical** | TRUE or FALSE only | `TRUE`, `FALSE` |

Use `class()` to ask R what type something is:

```{r}
# Numeric
streams <- 1500000000
class(streams)
```

```{r}
# Character
artist_name <- "Taylor Swift"
class(artist_name)
```

```{r}
# Logical
is_grammy_winner <- TRUE
class(is_grammy_winner)
```

::: callout-warning
## ️ Common Mistake: Forgetting Quotes

If you type `Taylor Swift` without quotes, R thinks you're referring to two variables named `Taylor` and `Swift` — which don't exist! You'll get an error.

Always wrap text in **double quotes** `" "` or **single quotes** `' '`.
:::

### Checking and Converting Types

```{r}
# is.numeric() asks "is this a number?" and returns TRUE or FALSE
is.numeric(streams)
```

```{r}
is.character(artist_name)
```

```{r}
# What if a number is stored as text?
fake_number <- "42"
class(fake_number)       # It's a character!
is.numeric(fake_number)  # FALSE — R can't do math with it
```

```{r}
# Convert it to a real number
real_number <- as.numeric(fake_number)
is.numeric(real_number)  # Now it's TRUE
```

------------------------------------------------------------------------

## Part 3: Working with Numbers

### Arithmetic Operators

```{r}
# Taylor Swift's Eras Tour had 152 shows in 2023-2024
# She played to an average of 72,000 fans per show
shows <- 152
avg_fans <- 72000

total_fans <- shows * avg_fans
total_fans
```

```{r}
# Format with commas using formatC (makes large numbers readable)
formatC(total_fans, format = "d", big.mark = ",")
```

::: callout-note
## R Arithmetic Operators

| Operator | Meaning        | Example   | Result |
|----------|----------------|-----------|--------|
| `+`      | Addition       | `5 + 3`   | `8`    |
| `-`      | Subtraction    | `10 - 4`  | `6`    |
| `*`      | Multiplication | `6 * 7`   | `42`   |
| `/`      | Division       | `20 / 4`  | `5`    |
| `^`      | Power          | `2^10`    | `1024` |
| `%%`     | Remainder      | `17 %% 5` | `2`    |
:::

```{r}
# Each Eras Tour ticket averaged $254
ticket_price <- 254
revenue_estimate <- total_fans * ticket_price
revenue_estimate

# Convert to billions
revenue_estimate / 1000000000
```

------------------------------------------------------------------------

## Part 4: Working with Text (Characters)

### Useful Functions for Text

```{r}
album <- "The Tortured Poets Department"

# How many characters long is this album title?
nchar(album)
```

```{r}
# Convert to uppercase
toupper(album)
```

```{r}
# Convert to lowercase
tolower(album)
```

```{r}
# Combine text with paste()
artist <- "Taylor Swift"
year   <- 2024

paste(artist, "released", album, "in", year)
```

```{r}
# paste0() combines with NO spaces
paste0("Track_", 1:5)   # We'll learn about 1:5 properly in the next lab!
```

::: callout-tip
## `paste()` vs `paste0()`

- `paste()` puts a **space** between items by default
- `paste0()` puts **nothing** between items

You can also change the separator in `paste()`: `paste("a", "b", "c", sep = "-")` gives `"a-b-c"`
:::

------------------------------------------------------------------------

## Part 5: Logical Values and Comparisons

Logical values (`TRUE` / `FALSE`) are the result of asking R a yes/no question. They are the foundation of filtering data — which you'll use constantly.

```{r}
# Comparison operators
midnights_sales <- 1600000   # first-week US sales
fearless_sales  <-  592000   # first-week US sales (original release)

midnights_sales > fearless_sales    # Is Midnights bigger?
```

```{r}
midnights_sales == fearless_sales   # Are they equal? (double == for comparison!)
```

```{r}
midnights_sales != fearless_sales   # Are they NOT equal?
```

::: callout-warning
## ️ `=` vs `==`

This is one of the most common beginner mistakes in R!

- `<-` or `=` — **assigns** a value to a variable
- `==` — **tests** whether two things are equal (asks a question)

`x <- 5` → puts 5 into x `x == 5` → asks "is x equal to 5?" and returns `TRUE` or `FALSE`
:::

```{r}
# Combining logical conditions
grammy_wins <- 14
billboards_number_ones <- 40

# AND: both must be true
grammy_wins > 10 & billboards_number_ones > 30
```

```{r}
# OR: at least one must be true
grammy_wins > 20 | billboards_number_ones > 30
```

------------------------------------------------------------------------

## Part 6: Useful Built-in Functions

R comes with hundreds of built-in functions. Here are a few you'll use all the time:

```{r}
# Some of Taylor Swift's Spotify monthly listener counts (in millions) over 6 months
listeners <- c(95.2, 98.7, 102.1, 89.4, 110.3, 107.6)
# (Don't worry about c() yet — we'll cover it properly in Lab 4!)

# Summary statistics
sum(listeners)       # Total
mean(listeners)      # Average
min(listeners)       # Lowest month
max(listeners)       # Highest month
range(listeners)     # Min and max together
length(listeners)    # How many values
```

```{r}
# round() controls decimal places
mean_listeners <- mean(listeners)
round(mean_listeners, 1)   # Round to 1 decimal place
```

::: callout-tip
## Getting Help in R

If you ever want to know what a function does or what arguments it takes, type `?` followed by the function name in the console:

``` r
?round
?mean
?paste
```

A help page will appear in the bottom-right panel. The **Examples** section at the bottom is usually the most useful part.
:::

------------------------------------------------------------------------

## Part 7: Putting It All Together

Let's write a small summary of Taylor Swift's career using everything we've learned:

```{r}
# --- Taylor Swift Career Summary ---

artist_name    <- "Taylor Swift"
debut_year     <- 2006
current_year   <- 2026
career_length  <- current_year - debut_year

studio_albums  <- 10
grammy_awards  <- 14
eras_revenue   <- 2077000000   # USD

# Print a summary
cat("Artist:", artist_name, "\n")
cat("Career length:", career_length, "years\n")
cat("Studio albums:", studio_albums, "\n")
cat("Grammy Awards:", grammy_awards, "\n")
cat("Eras Tour revenue: $", formatC(eras_revenue, format="d", big.mark=","), "\n")
cat("Is she a Grammy winner?", grammy_awards > 0, "\n")
```

::: callout-note
## What is `cat()`?

`cat()` ("concatenate and print") is used to print formatted output. `\n` means "new line." It's great for making readable summaries.
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together.*

::: callout-caution
## Knowledge Check Questions

**1.** What does `class("Fearless")` return?

```         
a) `"numeric"`  &nbsp;&nbsp; b) `"logical"`  &nbsp;&nbsp; c) `"character"`  &nbsp;&nbsp; d) `"text"`
```

**2.** What is the key difference between `=` and `==` in R?

**3.** Taylor Swift has 14 Grammy Awards. Which line of code correctly stores this?

```         
a) `grammys == 14`  &nbsp;&nbsp; b) `14 <- grammys`  &nbsp;&nbsp; c) `grammys <- 14`  &nbsp;&nbsp; d) `grammys = "14"`
```

**4.** What does `nchar("Midnights")` return?

```         
a) `1`  &nbsp;&nbsp; b) `9`  &nbsp;&nbsp; c) `"Midnights"`  &nbsp;&nbsp; d) `"character"`
```

**5.** True or False: `is.numeric("42")` returns `TRUE`.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** c) `"character"` — Any value in quotes is a character (text) type, regardless of what it says.

**2.** `=` (or `<-`) **assigns** a value to a variable. `==` **tests** whether two values are equal and returns `TRUE` or `FALSE`. This is one of the most common beginner mistakes.

**3.** c) `grammys <- 14` — Assignment goes left: the name on the left, the value on the right. Option d) is wrong because `"14"` is stored as text, not a number.

**4.** b) `9` — "Midnights" has 9 characters (M-i-d-n-i-g-h-t-s).

**5.** **False** — `"42"` is in quotes, so it is a **character**, not a numeric. `is.numeric("42")` returns `FALSE`. You would need `as.numeric("42")` to convert it first.
:::

------------------------------------------------------------------------

## Lab 2 Checklist

Before you leave, make sure you can:

- [ ] Create variables using `<-` and give them descriptive names
- [ ] Identify whether a value is `numeric`, `character`, or `logical`
- [ ] Use `class()` to check the type of a variable
- [ ] Do arithmetic with variables
- [ ] Use `paste()` to combine text
- [ ] Use comparison operators (`>`, `<`, `==`, `!=`) to get `TRUE`/`FALSE` results
- [ ] Use at least 3 of the summary functions: `sum()`, `mean()`, `min()`, `max()`, `length()`

::: callout-tip
## Bonus Challenge

Choose a different music artist. Create variables for their name, debut year, number of albums, and approximate total streams. Use `cat()` to print a formatted career summary. Add it to your Quarto document and render it!
:::

------------------------------------------------------------------------

## Before Next Class (Monday)

- Read **Intro2r §2.4.1–2.4.5, 3.1, 3.2, 7.2**
- Review your notes from today — especially the difference between `=` and `==`
- Remember: **Quiz 1 is this Friday** — it covers Labs 1 and 2

::: callout-important
## Quiz 1 Reminder

Quiz 1 is **in class this Friday (Aug 21)**. It will cover: - Basic R syntax and operators - Variable assignment - Data types (numeric, character, logical) - Simple functions (`class()`, `mean()`, `paste()`, etc.)

Review your notes and practice in the console!
:::

------------------------------------------------------------------------