---
title: "Lab 12: Data Visualization — Histograms, Bar Charts, Box Plots & Scatter Plots"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "September 21, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Seeing Data Clearly

A well-constructed graph communicates in seconds what a table of numbers takes minutes to reveal. In biology, the ability to visualise data is not a cosmetic skill — it is fundamental to understanding your results, spotting errors, and communicating findings to others.

This lab covers the four workhorses of scientific data visualisation using R's built-in graphics system: **histograms** for distributions, **bar charts** for categorical summaries, **box plots** for group comparisons, and **scatter plots** for relationships between two variables.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Choose the appropriate plot type for a given data structure
2. Create histograms with `hist()` and adjust bin width
3. Create bar charts with `barplot()` for categorical data
4. Create box plots with `boxplot()` to compare groups
5. Create scatter plots with `plot()` to explore relationships
6. Customise plots: axes, colours, labels, titles, and legends
7. Place multiple plots on one figure with `par(mfrow = ...)`
:::

------------------------------------------------------------------------

## Part 1: Choosing the Right Plot

Before writing any code, the most important question is: **what kind of data do I have?**

::: callout-important
## Plot Type Decision Guide

| Question | Data type | Best plot |
|----------|-----------|-----------|
| What does the distribution of one variable look like? | Continuous | Histogram |
| How do counts or totals compare across categories? | Categorical | Bar chart |
| How do distributions compare across groups? | Continuous + categorical | Box plot |
| Is there a relationship between two continuous variables? | Two continuous | Scatter plot |

The wrong plot type does not just look bad — it can actively mislead.
:::

We will use three biological datasets throughout this lab. Let us create them now:

```{r}
set.seed(42)

# Dataset 1: Enzyme activity (nmol/min/mg protein) across 4 temperature treatments
enzyme <- data.frame(
  temperature = rep(c("25C", "30C", "35C", "40C"), each = 20),
  activity    = c(rnorm(20, 18, 3),
                  rnorm(20, 26, 4),
                  rnorm(20, 31, 5),
                  rnorm(20, 22, 4))
)

# Dataset 2: Photosynthesis rate (umol CO2/m2/s) across 50 plant measurements
photo_rate <- c(rnorm(50, mean = 12.5, sd = 3.2))

# Dataset 3: Protein concentration vs absorbance (Bradford assay standard curve)
protein_conc <- seq(0, 2000, by = 200)     # ug/mL
absorbance   <- 0.00035 * protein_conc + rnorm(length(protein_conc), 0, 0.015)
```

------------------------------------------------------------------------

## Part 2: Histograms — Visualising Distributions

A histogram divides a continuous variable into bins and counts observations in each bin. It shows the **shape** of a distribution.

```{r}
# Basic histogram of photosynthesis rates
hist(photo_rate,
     xlab  = "Photosynthesis Rate (umol CO2/m2/s)",
     ylab  = "Frequency",
     main  = "Distribution of Photosynthesis Rates\n50 Plant Measurements",
     col   = "#2C5F8A",
     border = "white")
```

### Adjusting Bin Width

The number of bins dramatically affects what you see:

```{r}
# Compare different bin counts using par(mfrow)
par(mfrow = c(1, 3))   # 1 row, 3 columns of plots

hist(photo_rate, breaks = 5,  col = "#A8C4D9", border = "white",
     main = "5 bins",  xlab = "Rate")
hist(photo_rate, breaks = 15, col = "#2C5F8A", border = "white",
     main = "15 bins", xlab = "Rate")
hist(photo_rate, breaks = 30, col = "#1A3D5C", border = "white",
     main = "30 bins", xlab = "Rate")

par(mfrow = c(1, 1))   # Reset to single plot
```

::: callout-tip
## How Many Bins?

Too few bins hide the shape; too many create noise. A good rule of thumb: √n bins, where n is your sample size. For n = 50, try ~7–8 bins.

R's default (`breaks` not specified) usually makes a reasonable choice, but always inspect it.
:::

### Adding a Normal Curve

```{r}
# Histogram with a superimposed normal distribution curve
hist(photo_rate,
     freq   = FALSE,      # Switch y-axis to density (required for overlaying a curve)
     col    = "#A8C4D9",
     border = "white",
     xlab   = "Photosynthesis Rate (umol CO2/m2/s)",
     main   = "Photosynthesis Rate with Normal Curve")

# Overlay normal curve using estimated mean and SD
curve(dnorm(x, mean = mean(photo_rate), sd = sd(photo_rate)),
      add = TRUE, col = "#C8102E", lwd = 2)
```

------------------------------------------------------------------------

## Part 3: Bar Charts — Comparing Category Totals

Bar charts show a **summary statistic** (count, mean, total) for each category. They should not be used for raw continuous data — that is what box plots are for.

```{r}
# Mean enzyme activity at each temperature
mean_activity <- tapply(enzyme$activity, enzyme$temperature, mean)
mean_activity
```

```{r}
# Bar chart of mean enzyme activity
barplot(mean_activity,
        col    = c("#A8C4D9", "#5B8DB8", "#2C5F8A", "#1A3D5C"),
        ylab   = "Mean Enzyme Activity (nmol/min/mg protein)",
        xlab   = "Temperature",
        main   = "Mean Enzyme Activity at Four Temperatures",
        ylim   = c(0, 40))
```

### Adding Error Bars

A bar chart without error bars shows only the mean and conceals variability. Add standard error bars:

```{r}
# Calculate means and standard errors
means <- tapply(enzyme$activity, enzyme$temperature, mean)
sds   <- tapply(enzyme$activity, enzyme$temperature, sd)
ns    <- tapply(enzyme$activity, enzyme$temperature, length)
se    <- sds / sqrt(ns)

# Draw the bar chart and capture bar midpoint positions
bp <- barplot(means,
              col    = c("#A8C4D9", "#5B8DB8", "#2C5F8A", "#1A3D5C"),
              ylab   = "Mean Enzyme Activity (nmol/min/mg protein)",
              xlab   = "Temperature",
              main   = "Mean Enzyme Activity ± SE",
              ylim   = c(0, 42))

# Add error bars using arrows()
arrows(bp, means - se, bp, means + se,
       angle  = 90,
       code   = 3,
       length = 0.05,
       lwd    = 1.5)
```

::: callout-note
## Standard Deviation vs Standard Error

**Standard deviation (SD)**: describes the spread of your data — how variable individual measurements are.

**Standard error of the mean (SE = SD / √n)**: describes the precision of your estimate of the mean — how variable the mean would be if you repeated the experiment.

Use **SD** when describing your sample. Use **SE** (or a confidence interval) when making inferences about the population mean. Many published papers incorrectly use SE to make distributions look narrower than they are.
:::

------------------------------------------------------------------------

## Part 4: Box Plots — Comparing Distributions Across Groups

A box plot shows the **full distribution** of a continuous variable across groups: the median, interquartile range, and outliers. It is far more informative than a bar chart for this purpose.

```{r}
# Box plot of enzyme activity across all four temperature groups
boxplot(activity ~ temperature,
        data   = enzyme,
        col    = c("#A8C4D9", "#5B8DB8", "#2C5F8A", "#1A3D5C"),
        ylab   = "Enzyme Activity (nmol/min/mg protein)",
        xlab   = "Temperature",
        main   = "Enzyme Activity Distribution by Temperature",
        outline = TRUE)
```

::: callout-note
## Anatomy of a Box Plot

| Element | What it represents |
|---------|--------------------|
| Bottom of box | 25th percentile (Q1) |
| Line inside box | Median (50th percentile) |
| Top of box | 75th percentile (Q3) |
| Box height | Interquartile range (IQR = Q3 − Q1) |
| Whiskers | Extend to 1.5 × IQR from box edges |
| Points beyond whiskers | Potential outliers |
:::

### Adding Raw Data Points

Box plots can hide multi-modal distributions or small sample sizes. Always add the raw data when n is small:

```{r}
boxplot(activity ~ temperature,
        data    = enzyme,
        col     = "#A8C4D960",   # Transparent fill
        ylab    = "Enzyme Activity (nmol/min/mg protein)",
        xlab    = "Temperature",
        main    = "Enzyme Activity: Box Plot + Raw Data",
        outline = FALSE)

# Overlay individual data points with jitter
stripchart(activity ~ temperature,
           data     = enzyme,
           method   = "jitter",
           vertical = TRUE,
           add      = TRUE,
           pch      = 16,
           cex      = 0.7,
           col      = "#1A3D5C80")
```

------------------------------------------------------------------------

## Part 5: Scatter Plots — Exploring Relationships

Scatter plots show whether two continuous variables are related. This is essential for calibration curves, dose-response relationships, and regression analysis.

```{r}
# Bradford assay standard curve: absorbance vs protein concentration
plot(protein_conc, absorbance,
     pch  = 16,
     col  = "#2C5F8A",
     cex  = 1.2,
     xlab = "Protein Concentration (ug/mL)",
     ylab = "Absorbance (595 nm)",
     main = "Bradford Assay Standard Curve")

# Add a best-fit line
abline(lm(absorbance ~ protein_conc), col = "#C8102E", lwd = 2)

# Add a legend and correlation coefficient
r <- cor(protein_conc, absorbance)
legend("topleft",
       legend = paste0("r = ", round(r, 3)),
       bty    = "n")
```

::: callout-tip
## Interpreting a Standard Curve

The Bradford assay (and most spectrophotometric assays) produces a linear relationship between concentration and absorbance. Once you fit the line with `lm()`, you can rearrange it to calculate unknown protein concentrations from their absorbance readings.

We will cover linear regression in detail in Lab 22.
:::

------------------------------------------------------------------------

## Part 6: Combining Plots — `par(mfrow)`

```{r}
# Create a 2x2 panel showing all four plot types with the enzyme dataset
par(mfrow = c(2, 2),
    mar   = c(4, 4, 3, 1))   # Tighten margins

# Panel 1: Histogram of 35C activity
hist(enzyme$activity[enzyme$temperature == "35C"],
     col   = "#2C5F8A", border = "white",
     main  = "35C Activity Distribution",
     xlab  = "Activity")

# Panel 2: Bar chart of means
barplot(means,
        col   = c("#A8C4D9", "#5B8DB8", "#2C5F8A", "#1A3D5C"),
        main  = "Mean Activity by Temp",
        ylab  = "Mean activity")

# Panel 3: Box plot
boxplot(activity ~ temperature, data = enzyme,
        col  = "#5B8DB860",
        main = "Activity by Temperature",
        xlab = "Temperature", ylab = "Activity")

# Panel 4: Scatter of all data with jitter
plot(jitter(as.numeric(factor(enzyme$temperature))),
     enzyme$activity,
     col  = "#2C5F8A80",
     pch  = 16,
     xaxt = "n",
     xlab = "Temperature",
     ylab = "Activity",
     main = "Strip Plot")
axis(1, at = 1:4, labels = c("25C", "30C", "35C", "40C"))

par(mfrow = c(1, 1))   # Reset
```

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** A researcher measures serum glucose levels in 80 patients. Which plot is best for showing the distribution of these values?

    a) Bar chart    b) Scatter plot    c) Histogram    d) Pie chart

**2.** In a box plot, the line inside the box represents:

    a) The mean    b) The median    c) The mode    d) The standard deviation

**3.** You add error bars to a bar chart using `arrows()`. What should the bars represent if you are making inferences about the population mean?

    a) Range (min to max)    b) Standard deviation    c) Standard error or confidence interval    d) Variance

**4.** What does the R argument `freq = FALSE` in `hist()` change about the y-axis?

**5.** True or False: A box plot is more informative than a bar chart (mean ± SE) for visualising how a variable is distributed across groups.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** c) Histogram — A histogram is designed for showing the distribution (shape, spread, centre) of a single continuous variable. Bar charts are for categorical summaries.

**2.** b) The median — The centre line of a box plot always represents the **median** (50th percentile). The mean may be represented separately as a point. This is one reason box plots are robust to outliers — they are based on medians and percentiles.

**3.** c) Standard error or confidence interval — When making inferences about the true population mean, error bars should represent SE or CIs. SD describes sample variability, not estimation precision.

**4.** It switches the y-axis from **frequency (counts)** to **density** (proportion per unit). Density is required when overlaying a probability distribution curve, because the area under a density histogram equals 1, matching the area under a probability density function.

**5.** **True** — A bar chart (mean ± SE) shows only two numbers: the mean and a measure of precision. A box plot shows the median, IQR, whiskers, and outliers — giving a far richer picture of how data are distributed across groups. A bimodal distribution, for example, looks identical to a normal distribution in a mean ± SE bar chart.
:::

------------------------------------------------------------------------

## Lab 12 Checklist

Before you leave, make sure you can:

- [ ] Choose the correct plot type for a given biological question
- [ ] Create a histogram with `hist()` and adjust bin count with `breaks`
- [ ] Overlay a normal distribution curve on a density histogram
- [ ] Create a bar chart with `barplot()` and add standard error bars with `arrows()`
- [ ] Create a box plot with `boxplot()` using the formula interface `y ~ group`
- [ ] Overlay raw data points on a box plot using `stripchart()`
- [ ] Create a scatter plot with `plot()` and add a trend line with `abline(lm(...))`
- [ ] Arrange multiple plots with `par(mfrow = c(rows, cols))`

::: callout-tip
## Bonus Challenge

Using the `enzyme` data frame, produce a publication-quality box plot comparing enzyme activity across all four temperature groups. Add: (1) individual data points, (2) a horizontal reference line at the grand mean, (3) a meaningful title, and (4) axis labels with units. Export it as a PNG using `png("filename.png")` before and `dev.off()` after your plotting code.
:::

------------------------------------------------------------------------

## Before Next Class (Wednesday)

- Wednesday (Lab 13): ggplot2 — the publication-standard graphics system for R
- Read **R for Data Science Ch. 7–8** (link on Canvas)
- If you have not installed ggplot2 yet: open Posit Cloud, click on your project, and run `install.packages("ggplot2")` in the Console before class

------------------------------------------------------------------------
