---
title: "Lab 3: Generating Sequences of Numbers"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "August 24, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Sequences and the NBA

An 82-game NBA season generates an enormous amount of data — game numbers, jersey numbers, player ages, season years. In R, you don't have to type every number by hand. R gives you powerful tools to **generate sequences automatically**.

Today we'll learn three ways to create sequences: the `:` operator, `seq()`, and `rep()`.

::: callout-note
## Learning Objectives

By the end of this lab you will be able to:

1.  Use the `:` operator to create simple integer sequences
2.  Use `seq()` to create flexible sequences with custom spacing
3.  Use `rep()` to repeat values
4.  Combine sequences with `c()`
5.  Apply simple operations across an entire sequence at once
:::

------------------------------------------------------------------------

## Part 1: The `:` Operator — The Quick Sequence

The colon `:` is the fastest way to make a sequence of whole numbers.

```{r}
# NBA games are numbered 1 through 82 in a regular season
game_numbers <- 1:82
game_numbers
```

```{r}
# How many games are there?
length(game_numbers)
```

```{r}
# You can also go backwards
countdown <- 10:1
countdown
```

```{r}
# Sequences don't have to start at 1
jersey_numbers <- 23:32   # Numbers around LeBron James' famous #23
jersey_numbers
```

::: callout-tip
## When to Use `:`

Use `:` when you need consecutive whole numbers (1, 2, 3, ...).

It's great for: - Game numbers in a season - Row numbers in a dataset - Counting iterations
:::

------------------------------------------------------------------------

## Part 2: `seq()` — The Flexible Sequence

When you need more control — different step sizes, specific endpoints, or a set number of values — use `seq()`.

### `seq()` with a step size (`by =`)

```{r}
# Every 5th jersey number from 0 to 99 
every_fifth <- seq(from = 0, to = 99, by = 5)
every_fifth
```

```{r}
# Quarter times in a basketball game (12-minute quarters)
# Generate the minute marks at the end of each quarter
quarter_ends <- seq(from = 12, to = 48, by = 12)
quarter_ends
```

### `seq()` with a fixed number of values (`length.out =`)

```{r}
# Suppose we want exactly 10 evenly spaced points
# between a player's rookie salary ($1M) and max salary ($50M)
salary_steps <- seq(from = 1000000, to = 50000000, length.out = 10)
salary_steps

# Round to nearest dollar
round(salary_steps, 0)
```

::: callout-note
## `seq()` Arguments Summary

| Argument     | Meaning                | Example           |
|--------------|------------------------|-------------------|
| `from`       | Starting value         | `from = 0`        |
| `to`         | Ending value           | `to = 100`        |
| `by`         | Step between values    | `by = 5`          |
| `length.out` | Total number of values | `length.out = 10` |

You use **either** `by` **or** `length.out` — not both.
:::

```{r}
# NBA three-point line is 23.75 feet from the basket
# How many attempts from 15 to 30 feet in 0.5-foot increments?
distances <- seq(15, 30, by = 0.5)
distances
length(distances)
```

------------------------------------------------------------------------

## Part 3: `rep()` — Repeating Values

Sometimes you need to repeat the same value multiple times — like recording that a player played in 4 quarters, or that a team has 5 starters.

```{r}
# An NBA team has 5 starting players
# Record the starting designation for each player
starter_status <- rep("Starter", times = 5)
starter_status
```

```{r}
# A full roster has 15 players (5 starters + 10 bench)
bench_status <- rep("Bench", times = 10)
bench_status
```

```{r}
# Repeat a sequence of values
# Q1 through Q4, repeated across 3 games
quarters <- rep(1:4, times = 3)
quarters
```

```{r}
# What if each quarter is repeated to show multiple games?
quarter_labels <- rep(1:4, times = 5)   # 5 games
quarter_labels
```

### `rep()` with `each =`

```{r}
# 3 players, each listed twice (home game and away game)
players <- rep(c("LeBron", "Curry", "Durant"), each = 2)
players
```

::: callout-tip
## `times` vs `each`

- `rep(x, times = 3)` repeats the **whole thing** 3 times: `1 2 3 1 2 3 1 2 3`
- `rep(x, each = 3)` repeats **each element** 3 times: `1 1 1 2 2 2 3 3 3`
:::

```{r}
# Demonstrate the difference
rep(1:3, times = 3)
rep(1:3, each = 3)
```

------------------------------------------------------------------------

## Part 4: Combining with `c()`

`c()` stands for **combine** (or concatenate). It's how you stick multiple values or sequences together into one.

```{r}
# Stephen Curry's points per game in his last 8 games
curry_points <- c(28, 32, 19, 41, 25, 30, 22, 38)
curry_points
```

```{r}
# You can combine sequences too
first_half  <- 1:41
second_half <- 42:82
full_season <- c(first_half, second_half)
length(full_season)
```

```{r}
# Mix of types? R will convert everything to the most flexible type
mixed <- c(10, 20, "thirty")
mixed          # Notice the numbers got converted to text!
class(mixed)
```

::: callout-warning
## Type Coercion

When you mix types in `c()`, R automatically converts everything to the most general type. Numbers become text if there's any text in the mix. This is called **type coercion** and can cause unexpected results. Keep your vectors to one type when possible.
:::

------------------------------------------------------------------------

## Part 5: Math Across a Whole Sequence

One of R's superpowers: you can do math on **every element at once** without writing a loop.

```{r}
# LeBron James' points scored in 10 consecutive games
lebron_points <- c(26, 30, 19, 35, 28, 22, 31, 27, 33, 25)

# How many total points did he score?
sum(lebron_points)
```

```{r}
# What was his average?
mean(lebron_points)
```

```{r}
# Fantasy basketball often doubles points for star players
# Multiply every value by 2 at once!
fantasy_points <- lebron_points * 2
fantasy_points
```

```{r}
# Which games did he score 30 or more?
lebron_points >= 30
```

```{r}
# How many games did he score 30+?
sum(lebron_points >= 30)   # TRUE counts as 1, FALSE as 0
```

::: callout-note
## Vectorized Operations

In R, most operations automatically apply to **every element** in a sequence. This is called **vectorization** and it's what makes R so efficient — no need to write loops for basic math.

`lebron_points * 2` is equivalent to doing `26*2, 30*2, 19*2, ...` all at once.
:::

------------------------------------------------------------------------

## Part 6: Practical Example — An NBA Season Simulator

Let's put it all together. We'll build a simple season summary for the Golden State Warriors:

```{r}
# Game numbers for the full 82-game season
games <- 1:82

# Simulate points scored per game (we'll use real-ish averages)
# For now, let's create a simple pattern using sequences
# (In later labs we'll use real data!)

# Warriors tend to score between 110 and 130 points
# Let's create a realistic-looking sequence (not truly random yet)
set.seed(42)   # Makes our "random" numbers reproducible
warriors_points <- round(seq(110, 130, length.out = 82) +
                        rep(c(-8, 5, -3, 10, -6, 8, -4), length.out = 82))

# Season summary
cat("=== Golden State Warriors Season Summary ===\n")
cat("Total games:", length(games), "\n")
cat("Average points per game:", round(mean(warriors_points), 1), "\n")
cat("Highest scoring game:", max(warriors_points), "points\n")
cat("Lowest scoring game:", min(warriors_points), "points\n")
cat("Games scoring 120+:", sum(warriors_points >= 120), "\n")
```

```{r}
# Simple plot of points across the season
plot(games, warriors_points,
     type = "l",
     col  = "#1D428A",   # Warriors blue
     lwd  = 2,
     xlab = "Game Number",
     ylab = "Points Scored",
     main = "Golden State Warriors — Points Per Game\n2025–26 Season (simulated)")
abline(h = mean(warriors_points), col = "#FFC72C", lwd = 2, lty = 2)
legend("topright", legend = "Season Average", col = "#FFC72C", lty = 2, lwd = 2)
```

::: callout-tip
## Your First Plot!

Don't worry about understanding all the `plot()` arguments yet — we have a whole week dedicated to visualization. For now, just appreciate that R can draw a graph of your data in one function call.
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together.*

::: callout-caution
## Knowledge Check Questions

**1.** What does `5:9` produce in R?

    a) The number 59  &nbsp;&nbsp; b) `5 6 7 8 9`  &nbsp;&nbsp; c) `5 9`  &nbsp;&nbsp; d) An error

**2.** Which `seq()` argument controls how many values are generated?

    a) `from`  &nbsp;&nbsp; b) `to`  &nbsp;&nbsp; c) `by`  &nbsp;&nbsp; d) `length.out`

**3.** What is the output of `rep(c(1, 2), times = 3)`?

    a) `1 1 1 2 2 2`  &nbsp;&nbsp; b) `1 2 1 2 1 2`  &nbsp;&nbsp; c) `1 2 3 1 2 3`  &nbsp;&nbsp; d) `1 2`

**4.** What does `length(1:82)` return?

**5.** True or False: `sum(c(TRUE, FALSE, TRUE, TRUE))` returns `3`.

:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** b) `5 6 7 8 9` — The `:` operator creates a sequence of consecutive integers from the first number to the second, inclusive.

**2.** d) `length.out` — This argument specifies the exact number of evenly spaced values to generate between `from` and `to`.

**3.** b) `1 2 1 2 1 2` — `times = 3` repeats the **whole vector** 3 times. If you wanted `1 1 1 2 2 2`, you would use `each = 3` instead.

**4.** `82` — `1:82` creates a vector of 82 integers, and `length()` counts the number of elements.

**5.** **True** — In R, `TRUE` is treated as `1` and `FALSE` as `0` in arithmetic. So `sum(TRUE, FALSE, TRUE, TRUE)` = 1 + 0 + 1 + 1 = `3`.
:::

------------------------------------------------------------------------

## Lab 3 Checklist

Before you leave, make sure you can:

- [ ] Create a sequence using `:`
- [ ] Create a sequence using `seq()` with both `by =` and `length.out =`
- [ ] Use `rep()` with both `times =` and `each =`
- [ ] Combine values and sequences with `c()`
- [ ] Apply math operations to a whole sequence at once
- [ ] Use `sum()`, `mean()`, `min()`, `max()`, `length()` on a sequence

::: callout-tip
## Bonus Challenge

Create a sequence representing your favorite NBA player's jersey numbers worn throughout their career (look it up!). Use `rep()` to show how many seasons they wore each number. Calculate how many total seasons are represented.
:::

------------------------------------------------------------------------

## Before Next Class (Wednesday)

- Review `seq()`, `rep()`, and `c()` — write three examples of each in your notes
- Read **Intro2r §2.4.1–2.4.5**
- Wednesday (Lab 4) we build on this with **vectors and missing values**