---
title: "Lab 4: Vectors & Missing Values"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "August 26, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Vectors & Missing Data in the NFL

Real data is messy. Players get injured and miss games. Stats aren't always recorded. Measurement errors happen. In the NFL, a quarterback might play 14 of 17 games — the other 3 are `NA` (Not Available) in your dataset.

Today we learn how to build **vectors** properly and how to handle the inevitable **missing values** that appear in real biological and sports data.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Build and name vectors using `c()` and `names()`
2. Understand what `NA` means and why it matters
3. Use `is.na()` to detect missing values
4. Use `na.rm = TRUE` to calculate statistics despite missing data
5. Replace or remove `NA` values
6. Perform vector arithmetic
:::

------------------------------------------------------------------------

## Part 1: Building Vectors Properly

A **vector** is R's most fundamental data structure — a sequence of values all of the **same type**.

```{r}
# Patrick Mahomes' passing yards per game, 2024 season (first 10 games)
mahomes_yards <- c(267, 331, 198, 312, 279, 290, 341, 188, 255, 320)
mahomes_yards
```

```{r}
# Basic information about the vector
length(mahomes_yards)   # How many elements?
class(mahomes_yards)    # What type?
```

### Named Vectors

You can label each element in a vector, which makes your data much more readable:

```{r}
# Give each value a name corresponding to the opponent
opponents <- c("BAL", "CIN", "ATL", "LAC", "NO",
               "LV", "TB", "DEN", "CAR", "LV")

names(mahomes_yards) <- opponents
mahomes_yards
```

```{r}
# Now you can see which game each number belongs to
# Access by name
mahomes_yards["BAL"]
mahomes_yards["TB"]
```

::: callout-tip
## Why Name Your Vectors?
Names make vectors much easier to read and interpret. When you come back to your code weeks later, `mahomes_yards["BAL"]` is much clearer than `mahomes_yards[1]`.
:::

------------------------------------------------------------------------

## Part 2: Missing Values — `NA`

`NA` stands for **Not Available**. It is R's placeholder for missing data.

Think of it like a blank cell in a spreadsheet — the data should be there, but it isn't.

```{r}
# Lamar Jackson's rushing yards per game, 2024 season
# He missed 2 games due to injury (games 6 and 11)
jackson_rushing <- c(45, 62, 38, 71, 55, NA, 49, 83, 60, 41, NA, 77)
jackson_rushing
```

```{r}
# R tells you there are 2 NAs
summary(jackson_rushing)
```

### Detecting Missing Values with `is.na()`

```{r}
# Which positions are NA?
is.na(jackson_rushing)
```

```{r}
# How many games did he miss?
sum(is.na(jackson_rushing))
```

```{r}
# How many games did he play?
sum(!is.na(jackson_rushing))   # ! means "NOT"
```

::: callout-warning
## ️ NAs are Contagious!
If any value in your calculation is `NA`, the result is also `NA`:

```{r}
mean(jackson_rushing)   # Returns NA — because of the missing games!
```

This seems annoying, but it's actually protecting you. R is saying: "I don't know the answer because some data is missing."
:::

------------------------------------------------------------------------

## Part 3: Working Around Missing Values

### `na.rm = TRUE`

Most R statistical functions have an `na.rm` argument. Setting it to `TRUE` tells R to **remove the NAs before calculating**.

```{r}
# Now we get a real answer
mean(jackson_rushing, na.rm = TRUE)
```

```{r}
sum(jackson_rushing,  na.rm = TRUE)
max(jackson_rushing,  na.rm = TRUE)
min(jackson_rushing,  na.rm = TRUE)
```

::: callout-note
## What does `na.rm` stand for?
`na.rm` = **NA remove**. Setting it to `TRUE` removes `NA`s before doing the calculation.
:::

### Removing NAs Entirely

Sometimes you want a clean vector with no `NA`s at all:

```{r}
# na.omit() removes all NA values
jackson_clean <- na.omit(jackson_rushing)
jackson_clean
length(jackson_clean)   # 10 games (the 2 missed are gone)
```

```{r}
# Alternatively, use logical indexing (we'll cover this more in Lab 6)
jackson_played <- jackson_rushing[!is.na(jackson_rushing)]
jackson_played
```

### Replacing NAs with a Value

Sometimes you want to replace `NA` with something meaningful (like 0 for a missed game):

```{r}
jackson_with_zeros <- jackson_rushing
jackson_with_zeros[is.na(jackson_with_zeros)] <- 0
jackson_with_zeros
```

::: callout-important
## Should You Replace NAs with Zero?
It depends on the context! For rushing yards, replacing with 0 makes sense (if he didn't play, he gained 0 yards). But for body temperature or a test score, replacing missing data with 0 would be **wrong and misleading**.

In science, always document how you handled missing values and justify your choice.
:::

------------------------------------------------------------------------

## Part 4: Vector Arithmetic

Like we saw with sequences, arithmetic on vectors applies to **every element**:

```{r}
# Points scored by the Dallas Cowboys in each week of the season
cowboys_pts <- c(24, 17, 31, 20, 28, 14, 35, 22, 19, 27,
                 30, 21, 16, 28, 33, 18, 24)

# Points scored by opponents (what they allowed)
cowboys_allowed <- c(20, 21, 28, 17, 31, 20, 28, 30, 22, 14,
                     17, 24, 27, 21, 30, 27, 20)

# Point differential for each game (positive = win, negative = loss)
point_diff <- cowboys_pts - cowboys_allowed
point_diff
```

```{r}
# Which games did they win?
won_game <- point_diff > 0
won_game
```

```{r}
# How many wins?
sum(won_game)
```

```{r}
# How many losses?
sum(point_diff < 0)
```

```{r}
# Total points scored this season
sum(cowboys_pts)

# Average points per game
round(mean(cowboys_pts), 1)
```

### Arithmetic Between Two Vectors

```{r}
# Suppose each point is worth different fantasy values
# 1 point = 1 fantasy point, but bonus points for games over 28

bonus <- ifelse(cowboys_pts > 28, 3, 0)   # 3 bonus points for high-scoring games
fantasy_score <- cowboys_pts + bonus
fantasy_score
```

::: callout-tip
## `ifelse()` — Your First Conditional Function
`ifelse(test, value_if_TRUE, value_if_FALSE)` applies a test to every element and returns one of two values.

This is extremely useful and you'll use it constantly when cleaning data.
:::

------------------------------------------------------------------------

## Part 5: Putting It Together — QB Comparison

Let's compare two quarterbacks across a shared 10-game stretch, handling missing data properly:

```{r}
# 10-game passer rating comparison
qb_names <- c("Game 1", "Game 2", "Game 3", "Game 4", "Game 5",
               "Game 6", "Game 7", "Game 8", "Game 9", "Game 10")

# Patrick Mahomes passer ratings (0-158.3 scale)
mahomes_rating <- c(112.4, 98.7, 131.2, NA, 105.8,
                    122.1, 89.3, 141.5, 107.6, 118.9)

# Jalen Hurts passer ratings
hurts_rating   <- c(98.1, 107.5, 88.4, 125.3, 110.2,
                    NA, 119.7, 92.8, NA, 130.1)

# How many games did each miss?
cat("Mahomes missed:", sum(is.na(mahomes_rating)), "game(s)\n")
cat("Hurts missed:  ", sum(is.na(hurts_rating)), "game(s)\n")
```

```{r}
# Average passer rating (ignoring missed games)
mahomes_avg <- mean(mahomes_rating, na.rm = TRUE)
hurts_avg   <- mean(hurts_rating,   na.rm = TRUE)

cat("Mahomes average passer rating:", round(mahomes_avg, 1), "\n")
cat("Hurts average passer rating:  ", round(hurts_avg,   1), "\n")
cat("Difference:", round(mahomes_avg - hurts_avg, 1), "points\n")
```

```{r}
# Simple side-by-side bar chart
barplot(
  c(mahomes_avg, hurts_avg),
  names.arg = c("P. Mahomes", "J. Hurts"),
  col       = c("#E31837", "#004C54"),   # Chiefs red, Eagles midnight green
  ylim      = c(0, 160),
  ylab      = "Average Passer Rating",
  main      = "QB Passer Rating Comparison\n(missing games excluded)",
  border    = "white"
)
abline(h = 100, col = "gray50", lty = 2)
text(0.7, 102, "100 = Good QB Rating", col = "gray30", cex = 0.8)
```

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together.*

::: callout-caution
## Knowledge Check Questions

**1.** What does `NA` stand for in R?

    a) Not Applicable  &nbsp;&nbsp; b) Not Available  &nbsp;&nbsp; c) Null Answer  &nbsp;&nbsp; d) Negative Amount

**2.** You run `mean(c(10, 20, NA, 40))` and get `NA`. How do you fix this?

**3.** What does `sum(is.na(x))` calculate?

    a) The sum of all values in x  &nbsp;&nbsp; b) The number of missing values in x  &nbsp;&nbsp; c) Whether x has any NAs  &nbsp;&nbsp; d) The mean of x

**4.** True or False: `na.omit(x)` replaces NAs with zero.

**5.** You have a vector `scores <- c(88, NA, 91, 75, NA)`. Write the code to calculate the mean ignoring the missing values.

:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** b) Not Available — `NA` is R's way of representing missing or unknown data.

**2.** Add `na.rm = TRUE`: `mean(c(10, 20, NA, 40), na.rm = TRUE)` → returns `23.33`. The `na.rm` argument tells R to **remove NAs** before calculating.

**3.** b) The number of missing values in x — `is.na(x)` returns a logical vector of TRUE/FALSE, and `sum()` counts the TRUEs (each TRUE = 1).

**4.** **False** — `na.omit(x)` **removes** the NA values entirely, returning a shorter vector. To replace NAs with zero, you would use `x[is.na(x)] <- 0`.

**5.** `mean(scores, na.rm = TRUE)` → returns `84.67` (average of 88, 91, and 75).
:::

------------------------------------------------------------------------

## Lab 4 Checklist

Before you leave, make sure you can:

- [ ] Create a named vector using `c()` and `names()`
- [ ] Explain what `NA` means in R
- [ ] Use `is.na()` to find missing values
- [ ] Use `sum(is.na())` to count missing values
- [ ] Calculate `mean()`, `sum()`, etc. with `na.rm = TRUE`
- [ ] Use `na.omit()` to remove NAs
- [ ] Replace NAs using logical indexing `x[is.na(x)] <- value`
- [ ] Perform arithmetic between two vectors

::: callout-tip
## Bonus Challenge
Find the season statistics for two NFL quarterbacks (any two you like). Create named vectors for their touchdown passes in each week. One of them should have at least one `NA` for a missed game. Compare their averages using `na.rm = TRUE` and make a barplot. Add this to your Quarto document and render it.
:::

------------------------------------------------------------------------

## Before Next Class (Friday)

- Read **Intro2r §3.1, 3.2**
- Friday (Lab 5) we move to **data frames and matrices** — where rows and columns begin
- Practice: create a vector with at least one `NA` and calculate its mean both with and without `na.rm = TRUE`

------------------------------------------------------------------------
