---
title: "Lab 27: Measuring Species Diversity in R — Part II"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "November 2, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Comparing Communities and Understanding Composition

Lab 26 covered diversity *within* a single site. Now we ask: how do communities differ from each other, and how do we compare diversity fairly when sample sizes differ? This lab introduces rarefaction, diversity comparisons across sites, and non-metric multidimensional scaling (NMDS) — a technique for visualising community composition in two dimensions.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Explain why rarefaction is needed when sample sizes differ between sites
2. Produce and interpret rarefaction curves with `vegan::rarecurve()`
3. Compare Shannon diversity across sites using Kruskal-Wallis and pairwise tests
4. Explain the concept of beta diversity (turnover between sites)
5. Produce an NMDS ordination plot with `vegan::metaMDS()`
6. Overlay environmental gradients on an NMDS plot
:::

------------------------------------------------------------------------

## Part 1: The Sampling Problem — Why Rarefaction?

Comparing species richness between sites is only fair when sampling effort is equal. A site where you counted 1000 individuals will almost always appear richer than one where you counted 50 — not because it is more diverse, but because rare species are more likely to be detected with greater effort.

```{r}
# Demonstrate: richness increases with sample size
set.seed(42)

# Simulate a community with 30 species, log-normal abundance distribution
true_abundance <- round(rlnorm(30, meanlog = 3, sdlog = 1.2))
true_abundance <- sort(true_abundance, decreasing = TRUE)
species_pool   <- rep(paste0("Sp", 1:30), times = true_abundance)

# Draw samples of increasing size and count species found
sample_sizes <- c(10, 20, 50, 100, 200, 500, 1000)
richness_found <- sapply(sample_sizes, function(n) {
  samp <- sample(species_pool, min(n, length(species_pool)), replace = FALSE)
  length(unique(samp))
})

plot(sample_sizes, richness_found,
     type = "b", pch = 16, col = "#2C5F8A",
     xlab = "Number of Individuals Sampled",
     ylab = "Species Detected",
     main = "Species Accumulation Curve")
abline(h = 30, col = "#C8102E", lty = 2)
legend("bottomright", c("Observed", "True richness = 30"),
       col = c("#2C5F8A", "#C8102E"), lty = c(1, 2), bty = "n")
```

::: callout-important
## The Solution: Rarefaction

Rarefaction standardises sample sizes by asking: "How many species would we expect to find if we randomly drew only n individuals from this site?" It mathematically downsamples larger collections to a common sample size, making richness comparisons fair.

The rarefaction sample size is typically set to the smallest sample in the comparison.
:::

------------------------------------------------------------------------

## Part 2: Rarefaction Curves

```{r}
library(vegan)

# Simulate four forest sites with different sampling effort and true diversity
set.seed(7)
site_data <- matrix(
  c(rmultinom(1, 500, prob = runif(20)),   # Site 1: n=500, 20 spp
    rmultinom(1, 150, prob = runif(20)),   # Site 2: n=150, 20 spp (under-sampled)
    rmultinom(1, 400, prob = c(runif(12), rep(0, 8))),  # Site 3: n=400, only 12 spp
    rmultinom(1, 350, prob = runif(18))),  # Site 4: n=350, 18 spp
  nrow = 4, byrow = TRUE
)
colnames(site_data) <- paste0("Sp", 1:20)
rownames(site_data) <- paste0("Site_", 1:4)

# Raw species richness (unfair comparison)
specnumber(site_data)
```

```{r}
# Rarefaction curves — each curve shows expected richness at n individuals
rarecurve(site_data,
          step   = 10,
          col    = c("#2C5F8A", "#C8102E", "#5B8DB8", "#A8C4D9"),
          lwd    = 2,
          xlab   = "Number of Individuals",
          ylab   = "Expected Species Richness",
          main   = "Rarefaction Curves for Four Forest Sites",
          label  = TRUE)

# Vertical line at minimum sample size
abline(v = min(rowSums(site_data)), col = "grey40", lty = 2)
```

```{r}
# Rarefied richness at the minimum sample size
min_n <- min(rowSums(site_data))
cat("Minimum sample size:", min_n, "\n\n")
cat("Rarefied species richness at n =", min_n, ":\n")
print(rarefy(site_data, sample = min_n))
```

------------------------------------------------------------------------

## Part 3: Comparing Diversity Across Replicate Sites

To statistically compare diversity indices between habitat types, we need replicate plots within each habitat.

```{r}
set.seed(99)

# Simulate three habitat types with 8 replicate 1-ha plots each
# Habitat: pristine forest, disturbed forest, agricultural edge
generate_community <- function(n_plots, n_spp, total_ind, dominance_factor) {
  sapply(1:n_plots, function(i) {
    probs <- runif(n_spp)^dominance_factor   # Higher = more dominance
    rmultinom(1, total_ind, prob = probs / sum(probs))
  })
}

pristine   <- t(generate_community(8, 25, 300, 1))
disturbed  <- t(generate_community(8, 25, 300, 2.5))
agri_edge  <- t(generate_community(8, 25, 300, 4))

h_pristine  <- diversity(pristine,  "shannon")
h_disturbed <- diversity(disturbed, "shannon")
h_agri      <- diversity(agri_edge, "shannon")

diversity_df <- data.frame(
  H       = c(h_pristine, h_disturbed, h_agri),
  Habitat = rep(c("Pristine Forest", "Disturbed Forest", "Agricultural Edge"),
                each = 8)
)
diversity_df$Habitat <- factor(diversity_df$Habitat,
                                levels = c("Pristine Forest", "Disturbed Forest", "Agricultural Edge"))
```

```{r}
library(ggplot2)

ggplot(diversity_df, aes(x = Habitat, y = H, fill = Habitat)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.1, size = 2.5, alpha = 0.8) +
  scale_fill_manual(values = c("Pristine Forest"   = "#2C5F8A",
                               "Disturbed Forest"  = "#5B8DB8",
                               "Agricultural Edge" = "#C8102E")) +
  labs(title = "Shannon Diversity Across Habitat Types",
       x     = NULL, y = "Shannon H'") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

```{r}
# Kruskal-Wallis test (non-parametric — suitable for small n per group)
kruskal.test(H ~ Habitat, data = diversity_df)
```

```{r}
# Pairwise comparisons with Benjamini-Hochberg correction
pairwise.wilcox.test(diversity_df$H, diversity_df$Habitat,
                     p.adjust.method = "BH")
```

------------------------------------------------------------------------

## Part 4: Beta Diversity — Turnover Between Sites

**Alpha diversity** is diversity within a site. **Beta diversity** measures how much community composition changes between sites — i.e., turnover.

```{r}
# Bray-Curtis dissimilarity: 0 = identical communities, 1 = completely different
all_sites <- rbind(pristine, disturbed, agri_edge)
rownames(all_sites) <- c(paste0("Pris_", 1:8),
                         paste0("Dist_", 1:8),
                         paste0("Agri_", 1:8))

bray_dist <- vegdist(all_sites, method = "bray")

# Mean within-group and between-group dissimilarity
dist_matrix <- as.matrix(bray_dist)

cat("Mean Bray-Curtis dissimilarity:\n")
cat("  Within Pristine:    ",
    round(mean(dist_matrix[1:8, 1:8][upper.tri(dist_matrix[1:8, 1:8])]), 3), "\n")
cat("  Within Agricultural:",
    round(mean(dist_matrix[17:24, 17:24][upper.tri(dist_matrix[17:24, 17:24])]), 3), "\n")
cat("  Pristine vs Agri:   ",
    round(mean(dist_matrix[1:8, 17:24]), 3), "\n")
```

------------------------------------------------------------------------

## Part 5: NMDS Ordination

**Non-metric multidimensional scaling (NMDS)** takes a dissimilarity matrix and places communities in 2D space so that their distances on the plot reflect their dissimilarity in species composition. Closer points = more similar communities.

```{r}
# Run NMDS on the full community matrix
set.seed(42)
nmds_result <- metaMDS(all_sites,
                        distance = "bray",
                        k        = 2,   # 2 dimensions
                        trymax   = 50)

cat("NMDS stress:", round(nmds_result$stress, 3), "\n")
cat("Stress < 0.10 = good; < 0.20 = acceptable; > 0.20 = suspect\n")
```

```{r}
# Extract site scores (coordinates in ordination space)
scores_df <- as.data.frame(scores(nmds_result, display = "sites"))
scores_df$Habitat <- rep(c("Pristine Forest", "Disturbed Forest", "Agricultural Edge"),
                          each = 8)

ggplot(scores_df, aes(x = NMDS1, y = NMDS2, colour = Habitat, shape = Habitat)) +
  geom_point(size = 3.5, alpha = 0.9) +
  stat_ellipse(level = 0.80, linewidth = 0.8) +
  scale_colour_manual(values = c("Pristine Forest"   = "#2C5F8A",
                                 "Disturbed Forest"  = "#5B8DB8",
                                 "Agricultural Edge" = "#C8102E")) +
  labs(title    = "NMDS Ordination — Bird Community Composition",
       subtitle  = paste0("Bray-Curtis dissimilarity; stress = ",
                          round(nmds_result$stress, 3)),
       x        = "NMDS Axis 1",
       y        = "NMDS Axis 2") +
  theme_classic(base_size = 13)
```

::: callout-note
## Interpreting an NMDS Plot

- Axes have no inherent biological meaning — only the distances between points matter
- Points from the same habitat cluster together if communities are similar
- Widely separated clusters indicate distinct community composition (high beta diversity)
- Ellipses show 80% confidence regions for each group
- Low stress (< 0.10) means the 2D representation faithfully captures the true high-dimensional dissimilarity
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** You survey birds at two sites: Site A (n = 50 individuals, 12 species) and Site B (n = 200 individuals, 18 species). Why is it unfair to conclude that Site B is more species-rich?

**2.** A rarefaction curve for Site C continues to rise steeply even at the maximum sample size, while Site D's curve levels off completely. What does this tell you about each site?

**3.** On an NMDS plot with stress = 0.24, how should you interpret the ordination?

    a) The 2D plot faithfully represents community dissimilarity    b) The ordination should be trusted — 0.24 is excellent    c) The 2D representation is poor — consider 3D or different methods    d) The communities are more similar than they appear

**4.** You calculate Bray-Curtis dissimilarity between two plots and get 0.93. What does this mean?

**5.** True or False: Alpha diversity is a measure of community turnover between sites.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** Richness comparisons are only fair at equal sampling effort. Site B counted 4× more individuals, making it much more likely to detect rare species. By chance alone, sampling 200 individuals from the same community as Site A would yield more species than sampling 50. Rarefaction is needed: standardise both sites to n = 50 individuals and compare rarefied richness.

**2.** Site C's curve still rising = it has not been sampled adequately; there are likely more species present than detected. Sampling more individuals would reveal additional species. Site D's curve levelling off = the site is well-sampled; you have likely detected nearly all species present. Asymptotic curves indicate sampling is sufficient for richness estimation.

**3.** c) The 2D representation is poor — NMDS stress > 0.20 is generally considered suspect. The 2D plot cannot faithfully represent the true high-dimensional community dissimilarity. Options include: running NMDS in 3D (k = 3), trying different distance measures, or using alternative ordination methods (e.g., PCoA).

**4.** Bray-Curtis dissimilarity of 0.93 (on a 0–1 scale) indicates the two plots share almost no species in common — their communities are 93% dissimilar. Ecologically, this is extremely high turnover, suggesting the two plots are in very different environments or represent very different successional stages.

**5.** False — Alpha diversity is diversity **within** a site. **Beta diversity** measures turnover or compositional change **between** sites. Beta diversity can be quantified with dissimilarity indices (e.g., Bray-Curtis) or visualised with NMDS ordination.
:::

------------------------------------------------------------------------

## Lab 27 Checklist

Before you leave, make sure you can:

- [ ] Explain why rarefaction is needed for fair richness comparisons
- [ ] Produce rarefaction curves with `rarecurve()` and rarefied richness with `rarefy()`
- [ ] Compare Shannon H' across habitats with Kruskal-Wallis and pairwise Wilcoxon tests
- [ ] Calculate Bray-Curtis dissimilarity with `vegdist(method = "bray")`
- [ ] Run `metaMDS()` and check the stress value
- [ ] Produce a labelled NMDS scatter plot with `ggplot2` and interpret clustering

::: callout-tip
## Bonus Challenge

Using the `vegan` BCI dataset (50 plots, tropical tree species), run an NMDS ordination. Overlay species scores on the plot using `scores(nmds, "species")` — which species appear to drive the differences between plots? Are there any obvious gradients? As a challenge, use `envfit()` to test whether plot coordinates (plot number) correlate significantly with the ordination axes.
:::

------------------------------------------------------------------------

## Before Next Class (Wednesday)

- Wednesday (Lab 28): Building interactive web apps in R with Shiny
- Friday (Lab 29): Dose-response curves and IC50 calculation in pharmacology
- **Quiz 7** is Monday of Week 13 — covers Labs 25–29

------------------------------------------------------------------------
