---
title: "Lab 7: Dates & Times; The Apply Family"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "September 2, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Dates, Times & the Apply Family — with MLB Data

Baseball has more data than almost any other sport — games played every day for six months, timestamps on every pitch, player performance tracked by date. Today we learn how to work with **dates and times** in R, and how to apply functions efficiently across rows, columns, or groups using the **apply family**.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Create and format dates using `as.Date()` and `format()`
2. Calculate time differences between dates
3. Extract components from dates (month, year, day of week)
4. Use `apply()` on matrices
5. Use `lapply()` and `sapply()` on lists and vectors
6. Use `tapply()` to calculate group summaries
:::

------------------------------------------------------------------------

## Part 1: Working with Dates

In R, dates are a special data type. They look like text but behave like numbers — you can subtract them, compare them, and extract parts.

### Creating Dates with `as.Date()`

```{r}
# The 2026 MLB season Opening Day
opening_day <- as.Date("2026-03-26")
opening_day
class(opening_day)
```

::: callout-important
## The Standard Date Format in R
R expects dates in **YYYY-MM-DD** format (year-month-day) by default. This is called ISO 8601 format.

-  `"2026-09-01"` — works perfectly
-  `"Sept 1, 2026"` — R won't understand this without extra instructions
-  `"09/01/2026"` — you need to specify the format

If your date is in a different format, use the `format` argument:
```r
as.Date("09/01/2026", format = "%m/%d/%Y")
```
:::

```{r}
# Key MLB 2026 season dates
spring_training <- as.Date("2026-02-14")
opening_day     <- as.Date("2026-03-26")
all_star_game   <- as.Date("2026-07-14")
playoff_start   <- as.Date("2026-10-01")
world_series    <- as.Date("2026-10-21")
```

### Calculating Time Differences

```{r}
# How many days from Opening Day to the All-Star Game?
all_star_game - opening_day
```

```{r}
# How many days is the regular season?
as.numeric(playoff_start - opening_day)
```

```{r}
# Days between Spring Training and World Series
as.numeric(world_series - spring_training)
```

::: callout-tip
## `as.numeric()` on Date Differences
When you subtract two dates, R returns a "difftime" object. Wrapping it in `as.numeric()` gives you a plain number (in days) that you can use in calculations.
:::

### Today's Date

```{r}
# Sys.Date() gives today's date automatically
today <- Sys.Date()
today

# How many days until the World Series?
as.numeric(world_series - today)
```

------------------------------------------------------------------------

## Part 2: Formatting and Extracting Date Components

The `format()` function lets you display dates in any way you like, and extract specific components:

```{r}
# Display opening day in a friendly format
format(opening_day, "%B %d, %Y")      # "March 26, 2026"
```

```{r}
format(opening_day, "%m/%d/%Y")       # "03/26/2026"
```

```{r}
format(opening_day, "%A")             # Day of the week
```

::: callout-note
## Date Format Codes

| Code | Meaning | Example |
|------|---------|---------|
| `%Y` | 4-digit year | 2026 |
| `%y` | 2-digit year | 26 |
| `%m` | Month (number) | 03 |
| `%B` | Month (full name) | March |
| `%b` | Month (abbreviated) | Mar |
| `%d` | Day of month | 26 |
| `%A` | Day of week (full) | Thursday |
| `%a` | Day of week (abbreviated) | Thu |
:::

```{r}
# Extract just the month number
as.numeric(format(opening_day, "%m"))    # 3 = March
```

```{r}
# What month is the World Series?
format(world_series, "%B %Y")
```

------------------------------------------------------------------------

## Part 3: Working with Sequences of Dates

```{r}
# Generate all game dates for the first week of the 2026 MLB season
first_week <- seq(from = opening_day,
                  to   = opening_day + 6,
                  by   = "day")
first_week
```

```{r}
# Show day names for the first week
format(first_week, "%A %b %d")
```

```{r}
# Generate monthly milestones through the season
monthly <- seq(from = opening_day,
               to   = playoff_start,
               by   = "month")
format(monthly, "%B %d, %Y")
```

------------------------------------------------------------------------

## Part 4: The Apply Family

The apply family of functions lets you apply the same function to many things at once — without writing a repetitive loop. This is very important in biology when you have many samples, genes, or games to process.

### `apply()` — For Matrices (Rows or Columns)

```{r}
# Season stats matrix for 5 Dodgers players
# Columns: Batting Average, Home Runs, RBIs, Stolen Bases
dodgers_stats <- matrix(
  c(0.312, 32, 95, 8,
    0.285, 28, 87, 22,
    0.301, 19, 74, 35,
    0.267, 41, 108, 3,
    0.320, 12, 58, 44),
  nrow  = 5,
  byrow = TRUE
)
rownames(dodgers_stats) <- c("Betts", "Freeman", "Hernandez", "Muncy", "Outman")
colnames(dodgers_stats) <- c("AVG", "HR", "RBI", "SB")

dodgers_stats
```

```{r}
# apply() syntax: apply(matrix, MARGIN, function)
# MARGIN = 1 means apply to each ROW
# MARGIN = 2 means apply to each COLUMN

# Average value for each stat (column averages)
apply(dodgers_stats, 2, mean)
```

```{r}
# Total for each player across all stats (not meaningful here, but shows the concept)
apply(dodgers_stats, 1, sum)
```

```{r}
# Which player has the highest value in each stat?
apply(dodgers_stats, 2, which.max)
```

::: callout-tip
## Remember: MARGIN = 1 (Rows) or 2 (Columns)?
A handy way to remember:
- **1** = rows → think of the **1st dimension**
- **2** = columns → think of the **2nd dimension**

Or: "1 goes across, 2 goes down"
:::

------------------------------------------------------------------------

### `sapply()` — Apply to a Vector or List, Get Simple Output

`sapply()` applies a function to each element of a vector or list and tries to return a simple result (a vector or matrix):

```{r}
# Player names vector
players <- c("Mookie Betts", "Freddie Freeman", "Miguel Rojas",
              "Max Muncy", "James Outman")

# Apply nchar() to get the length of each player's name
sapply(players, nchar)
```

```{r}
# Calculate the square root of each HR total
hr_totals <- c(32, 28, 19, 41, 12)
sapply(hr_totals, sqrt)
```

```{r}
# Custom function: categorize HR totals
hr_category <- function(hr) {
  if (hr >= 35) return("Power Hitter")
  if (hr >= 20) return("Average Power")
  return("Contact Hitter")
}

sapply(hr_totals, hr_category)
```

::: callout-note
## Writing Your Own Functions
You just saw R let you define your own function with `function() { }`. We'll go deeper into custom functions in future labs, but notice how natural the syntax is:

```r
my_function <- function(input) {
  # do something with input
  return(result)
}
```
:::

------------------------------------------------------------------------

### `lapply()` — Apply to a List, Always Return a List

`lapply()` is like `sapply()` but always returns a **list** (useful when results have different lengths):

```{r}
# A list of game scores for 3 Dodgers games
game_scores <- list(
  game_1 = c(7, 2, 4, 0, 3, 1, 2, 0, 0),   # Runs per inning
  game_2 = c(0, 0, 3, 5, 0, 2, 0, 1, 2),
  game_3 = c(1, 1, 0, 0, 4, 0, 3, 2, 1)
)

# Calculate total runs in each game
lapply(game_scores, sum)
```

```{r}
# Use sapply to get a simpler vector output
sapply(game_scores, sum)
```

```{r}
# Which inning did they score the most runs in each game?
sapply(game_scores, which.max)
```

------------------------------------------------------------------------

### `tapply()` — Apply by Group

`tapply()` is extremely useful in biology and sports: it applies a function to a variable **split by a grouping variable**.

```{r}
# MLB player data: batting average by position
mlb_players <- data.frame(
  player   = c("Betts", "Freeman", "Turner", "Muncy",
                "Smith", "Outman", "Hernandez", "Barnes", "Buehler"),
  position = c("OF", "1B", "SS", "3B",
                "C", "OF", "2B", "C", "SP"),
  avg      = c(0.312, 0.285, 0.271, 0.267,
                0.250, 0.320, 0.301, 0.239, 0.098),
  hr       = c(32, 28, 15, 41, 18, 12, 19, 11, 2)
)

# Average batting average by position
tapply(mlb_players$avg, mlb_players$position, mean)
```

```{r}
# Total home runs by position
tapply(mlb_players$hr, mlb_players$position, sum)
```

::: callout-tip
## `tapply()` in Biology
`tapply()` is exactly what you use to compare groups in biology — e.g., average gene expression in treatment vs. control, or species richness in disturbed vs. undisturbed habitats. This pattern is at the heart of almost all biological data analysis.
:::

------------------------------------------------------------------------

## Part 5: Putting It All Together

Let's analyze a simulated MLB pitch dataset combining dates and apply functions:

```{r}
# Simulated pitching data for 3 Dodgers starting pitchers over 6 starts each
set.seed(123)
pitching <- data.frame(
  pitcher     = rep(c("Buehler", "Kershaw", "May"), each = 6),
  start_date  = as.Date(c("2026-04-05","2026-04-11","2026-04-17",
                           "2026-04-23","2026-04-29","2026-05-05",
                           "2026-04-06","2026-04-12","2026-04-18",
                           "2026-04-24","2026-04-30","2026-05-06",
                           "2026-04-07","2026-04-13","2026-04-19",
                           "2026-04-25","2026-05-01","2026-05-07")),
  innings     = c(6.0, 5.2, 7.0, 6.1, 5.0, 6.2,
                  7.0, 7.1, 6.0, 8.0, 5.2, 7.0,
                  5.0, 6.0, 5.1, 6.2, 4.1, 5.0),
  earned_runs = c(3, 4, 1, 2, 5, 3, 2, 1, 3, 0, 4, 2, 4, 3, 5, 2, 6, 4)
)

# Calculate ERA per start: ERA = (earned_runs / innings) * 9
pitching$era_per_start <- round((pitching$earned_runs / pitching$innings) * 9, 2)

# Average ERA by pitcher using tapply
cat("Average ERA by Pitcher:\n")
tapply(pitching$era_per_start, pitching$pitcher, mean)
```

```{r}
# Which month do most starts occur in?
pitching$month <- format(pitching$start_date, "%B")
table(pitching$month)
```

```{r}
# Visualize ERA by pitcher
boxplot(era_per_start ~ pitcher, data = pitching,
        col  = c("#005A9C", "#EF3E42", "#A5ACAF"),  # Dodger colors
        ylab = "ERA per Start",
        xlab = "Pitcher",
        main = "Starting Pitcher ERA Distribution\nLos Angeles Dodgers, Spring 2026")
abline(h = 4.50, col = "red", lty = 2)
text(0.5, 4.7, "4.50 ERA threshold", col = "red", cex = 0.8, adj = 0)
```

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together.*

::: callout-caution
## Knowledge Check Questions

**1.** Which of these correctly creates a date in R?

    a) `as.Date("July 4, 2026")`  &nbsp;&nbsp; b) `as.Date("2026-07-04")`  &nbsp;&nbsp; c) `date("04/07/2026")`  &nbsp;&nbsp; d) `"2026-07-04"`

**2.** You subtract two date objects: `as.Date("2026-10-31") - as.Date("2026-09-01")`. What type of result does R return, and how would you convert it to a plain number?

**3.** In `apply(my_matrix, 2, mean)`, what does the `2` refer to?

    a) Apply to every 2nd row  &nbsp;&nbsp; b) Apply to columns  &nbsp;&nbsp; c) Apply to rows  &nbsp;&nbsp; d) Apply the function twice

**4.** What is the main difference between `sapply()` and `lapply()`?

**5.** Which function from the apply family would you use to calculate the average score **separately for each team** in a data frame?

    a) `apply()`  &nbsp;&nbsp; b) `lapply()`  &nbsp;&nbsp; c) `sapply()`  &nbsp;&nbsp; d) `tapply()`

:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** b) `as.Date("2026-07-04")` — R's default date format is **YYYY-MM-DD** (ISO 8601). Other formats require you to specify `format = "%m/%d/%Y"` etc.

**2.** R returns a **difftime** object. Wrap it in `as.numeric()` to get a plain number of days: `as.numeric(as.Date("2026-10-31") - as.Date("2026-09-01"))` → `60`.

**3.** b) Apply to columns — In `apply()`, MARGIN = 1 applies to **rows**, MARGIN = 2 applies to **columns**. Here, `mean` is calculated for each column of the matrix.

**4.** `sapply()` tries to simplify the output to a **vector or matrix**. `lapply()` always returns a **list**. Use `sapply()` when you want a clean vector; use `lapply()` when results might have different lengths.

**5.** d) `tapply()` — This function applies a function to a variable **split by a grouping variable**, making it ideal for group-wise summaries like average score per team.
:::

------------------------------------------------------------------------

## Lab 7 Checklist

Before you leave, make sure you can:

- [ ] Create a date with `as.Date()` in YYYY-MM-DD format
- [ ] Calculate the number of days between two dates
- [ ] Use `format()` to display dates in different formats
- [ ] Extract date components (month, day, weekday)
- [ ] Use `apply()` with MARGIN = 1 (rows) and 2 (columns)
- [ ] Use `sapply()` to apply a function to a vector
- [ ] Use `lapply()` when the output is a list
- [ ] Use `tapply()` to get group summaries

::: callout-tip
## Bonus Challenge
Create a data frame of any sports team's game schedule with at least 10 dates. Use `format()` to find the day of the week for each game. Use `tapply()` to calculate the average score on weekday games vs. weekend games (hint: Saturday and Sunday are weekends).
:::

------------------------------------------------------------------------

## Before Next Class 

- Monday is **Labor Day — No Class** ️
- **Wednesday Sep 9** (Lab 8): Hypothesis Testing & the Null Hypothesis
- Read **Intro2r Ch. 4** 
- Remember: **Quiz 2 is Friday Sep 4** — covers Labs 3–7

::: callout-important
## Quiz 2 — This Friday!
Topics covered: sequences, vectors, missing values, data frames, matrices, subsetting, dates, and apply functions.

Best way to prepare: open a new Quarto document and try to reproduce the key examples from Labs 3–7 from memory.
:::

