---
title: "Lab 10: Frequency Data, Confidence Intervals for Proportions & the Poisson Distribution"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "September 14, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Counting and Categorising: Frequency Data in Biology

Much of the data collected in biology is not a measurement on a continuous scale but a **count** or a **category**. How many patients responded to a treatment? What proportion of offspring show a recessive phenotype? How many mutations occur per genome replication? These are all questions about **frequency data**.

In this lab we cover three connected ideas: summarising categorical data with frequency tables, estimating proportions with confidence intervals, and modelling rare biological events with the Poisson distribution.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Create frequency tables with `table()` and `prop.table()`
2. Visualise frequency data with bar charts
3. Compute a confidence interval for a proportion using `prop.test()`
4. Explain what the Poisson distribution models
5. Use `dpois()`, `ppois()`, and `rpois()` to work with Poisson probabilities
6. Recognise which biological questions call for Poisson modelling
:::

------------------------------------------------------------------------

## Part 1: Frequency Tables — ABO Blood Types

The ABO blood group system is determined by alleles at a single locus and follows predictable population frequencies. In the United States the approximate frequencies are: O = 44%, A = 42%, B = 10%, AB = 4%.

Suppose you blood-type 200 students at a university health clinic:

```{r}
# Simulated blood type data from 200 students
set.seed(42)
blood_types <- sample(
  x       = c("O", "A", "B", "AB"),
  size    = 200,
  replace = TRUE,
  prob    = c(0.44, 0.42, 0.10, 0.04)
)

# First six observations
head(blood_types, 20)
```

### Counts with `table()`

```{r}
# Count how many students fall into each blood type
bt_counts <- table(blood_types)
bt_counts
```

```{r}
# Sort from most to least common
sort(bt_counts, decreasing = TRUE)
```

### Proportions with `prop.table()`

```{r}
# Convert counts to proportions (values between 0 and 1)
bt_props <- prop.table(bt_counts)
round(bt_props, 3)
```

```{r}
# Convert to percentages
round(bt_props * 100, 1)
```

::: callout-tip
## `table()` vs `prop.table()`

`table()` gives you **raw counts** — how many of each category.

`prop.table()` takes a table and converts counts to **proportions** that sum to 1.

Use `prop.table(table(x))` when you want to compare across groups of different sizes.
:::

### Visualising with a Bar Chart

```{r}
# Order by frequency for a cleaner plot
bt_ordered <- sort(bt_counts, decreasing = TRUE)

barplot(bt_ordered,
        col    = c("#2C5F8A", "#5B8DB8", "#A8C4D9", "#D6E8F5"),
        ylab   = "Number of Students",
        xlab   = "Blood Type",
        main   = "ABO Blood Type Distribution\n200 University Students (Simulated)",
        ylim   = c(0, 120))

# Add reference line at expected US frequency
abline(h = 200 * 0.44, col = "red", lty = 2)
text(4.5, 200 * 0.44 + 3, "Expected O frequency", col = "red", cex = 0.8)
```

------------------------------------------------------------------------

## Part 2: Two-Way Frequency Tables

We can cross-tabulate two categorical variables at once. Suppose we also recorded whether each student is Rh-positive or Rh-negative (approximately 85% of people are Rh+):

```{r}
set.seed(7)
rh_factor <- sample(c("Rh+", "Rh-"), size = 200, replace = TRUE, prob = c(0.85, 0.15))

# Two-way table: blood type by Rh factor
two_way <- table(blood_types, rh_factor)
two_way
```

```{r}
# Row proportions: within each blood type, what fraction is Rh+?
round(prop.table(two_way, margin = 1), 3)
```

::: callout-note
## `margin` in `prop.table()`

- `margin = 1` → proportions calculated **within each row** (row sums to 1)
- `margin = 2` → proportions calculated **within each column** (column sums to 1)
- No `margin` argument → all cells sum to 1 (overall proportions)
:::

```{r}
# Visualise as a grouped bar chart
barplot(t(two_way),
        beside  = TRUE,
        col     = c("#2C5F8A", "#C8102E"),
        legend  = rownames(t(two_way)),
        xlab    = "Blood Type",
        ylab    = "Count",
        main    = "ABO Blood Type by Rh Factor")
```

------------------------------------------------------------------------

## Part 3: Confidence Intervals for Proportions

A frequency or proportion from a sample is just an **estimate** of the true population proportion. We need a confidence interval to express our uncertainty.

### The `prop.test()` Function

Suppose in our sample, 88 out of 200 students have blood type A. We want to estimate the true proportion of type-A individuals in the population with a 95% confidence interval.

```{r}
# 88 successes (type A) out of 200 students
result <- prop.test(x = 88, n = 200, conf.level = 0.95)
result
```

```{r}
# Extract just the confidence interval
result$conf.int
```

```{r}
# Extract the point estimate (sample proportion)
result$estimate
```

::: callout-note
## Reading `prop.test()` Output

| Output element | Meaning |
|----------------|---------|
| `X-squared` | Chi-square statistic |
| `p-value` | Tests H₀: p = 0.5 by default |
| `95 percent confidence interval` | The range the true proportion likely falls in |
| `sample estimates: p` | Your observed proportion |

The confidence interval is what you usually want. For a two-sided 95% CI, you can be 95% confident the true proportion lies within those bounds.
:::

### Testing Against a Known Proportion

The US population frequency of blood type A is 0.42. Is our sample consistent with this?

```{r}
# H₀: true proportion of type A = 0.42
# H₁: true proportion ≠ 0.42
prop.test(x = 88, n = 200, p = 0.42, conf.level = 0.95)
```

```{r}
# Visualise: observed proportion with CI
obs_prop <- 88 / 200
ci       <- prop.test(88, 200)$conf.int

plot(x    = 1,
     y    = obs_prop,
     xlim = c(0.5, 1.5),
     ylim = c(0.30, 0.60),
     pch  = 16, cex = 1.5,
     col  = "#2C5F8A",
     xaxt = "n",
     xlab = "",
     ylab = "Proportion",
     main = "Observed Proportion of Blood Type A\nwith 95% Confidence Interval")
arrows(1, ci[1], 1, ci[2], angle = 90, code = 3, length = 0.1, col = "#2C5F8A", lwd = 2)
abline(h = 0.42, col = "red", lty = 2)
text(1.2, 0.42, "US expected (0.42)", col = "red", cex = 0.8)
```

------------------------------------------------------------------------

## Part 4: The Poisson Distribution — Modelling Rare Biological Events

The **Poisson distribution** describes the probability of a given number of events occurring in a fixed interval of time or space, when those events happen at a known average rate and are independent of each other.

In biology, it models:

- The number of mutations per genome replication
- The number of ion channel openings per second
- The number of cells in a haemocytometer grid square
- The number of rare disease cases per 100,000 people per year

::: callout-important
## The Key Parameter: Lambda (λ)

The Poisson distribution has a single parameter, **λ** (lambda) = the mean number of events per interval.

If λ = 2 mutations per genome replication, then the Poisson distribution tells you the probability of seeing 0, 1, 2, 3... mutations in any given replication event.

Crucially: **the mean equals the variance** in a Poisson distribution. This is a useful diagnostic check — if your data's variance is much larger than its mean, the Poisson may not be the right model.
:::

### Poisson Probability Functions

```{r}
# Suppose the spontaneous mutation rate is 2 mutations per genome replication (lambda = 2)
lambda <- 2

# P(exactly 0 mutations)
dpois(0, lambda = lambda)
```

```{r}
# P(exactly 3 mutations)
dpois(3, lambda = lambda)
```

```{r}
# Full probability distribution: P(0 through 10 mutations)
k        <- 0:10
probs    <- dpois(k, lambda = lambda)
names(probs) <- k
round(probs, 4)
```

```{r}
# Visualise the distribution
barplot(probs,
        names.arg = k,
        col       = "#2C5F8A",
        xlab      = "Number of Mutations per Replication",
        ylab      = "Probability",
        main      = paste0("Poisson Distribution (lambda = ", lambda, ")\nMutation Rate Model"))
```

### Cumulative Probabilities with `ppois()`

```{r}
# P(3 or fewer mutations) — cumulative
ppois(3, lambda = lambda)
```

```{r}
# P(more than 5 mutations) — upper tail
ppois(5, lambda = lambda, lower.tail = FALSE)
```

::: callout-tip
## `dpois()` vs `ppois()`

- `dpois(k, lambda)` — probability of **exactly** k events
- `ppois(k, lambda)` — probability of **k or fewer** events (cumulative)
- `ppois(k, lambda, lower.tail = FALSE)` — probability of **more than** k events

The same logic applies to `dnorm`/`pnorm` and other distribution functions in R.
:::

### Simulating Poisson Data

```{r}
# Simulate 1000 genome replications, each with lambda = 2 mutations
set.seed(123)
simulated_mutations <- rpois(1000, lambda = 2)

cat("Mean mutations per replication:    ", round(mean(simulated_mutations), 3), "\n")
cat("Variance of mutations:             ", round(var(simulated_mutations), 3), "\n")
cat("(For Poisson, mean ≈ variance)\n")
```

```{r}
# Compare simulated data to theoretical Poisson
obs_freq   <- table(simulated_mutations) / 1000
theory_prob <- dpois(as.numeric(names(obs_freq)), lambda = 2)

barplot(rbind(obs_freq, theory_prob),
        beside    = TRUE,
        col       = c("#2C5F8A", "#C8102E"),
        legend    = c("Simulated", "Theoretical Poisson"),
        xlab      = "Number of Mutations",
        ylab      = "Proportion",
        main      = "Simulated vs Theoretical Poisson(lambda=2)")
```

### Real Application: Rare Disease Incidence

A disease occurs at a rate of 3.5 cases per 100,000 people per year. A city of 500,000 people is monitored. On average, how many cases do we expect per year, and what is the probability of seeing 25 or more cases?

```{r}
# Expected cases in a city of 500,000
lambda_city <- 3.5 * (500000 / 100000)
cat("Expected cases per year:", lambda_city, "\n")
```

```{r}
# Probability of 25 or more cases
p_25_plus <- ppois(24, lambda = lambda_city, lower.tail = FALSE)
cat("P(25 or more cases):", round(p_25_plus, 4), "\n")
```

```{r}
# Plot the distribution across plausible outcomes
k_range <- 0:35
plot(k_range, dpois(k_range, lambda = lambda_city),
     type = "h",
     lwd  = 3,
     col  = ifelse(k_range >= 25, "#C8102E", "#2C5F8A"),
     xlab = "Number of Cases per Year",
     ylab = "Probability",
     main = "Poisson Distribution: Rare Disease Cases\nCity of 500,000 (lambda = 17.5)")
legend("topright",
       legend = c("Expected range", "25+ cases (unusual)"),
       col    = c("#2C5F8A", "#C8102E"),
       lty    = 1, lwd = 3)
```

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** You survey 150 patients and find 45 have hypertension. What R function would you use to calculate a 95% confidence interval for the proportion of hypertensive patients?

    a) `t.test(45, 150)`    b) `prop.test(45, 150)`    c) `table(45, 150)`    d) `pnorm(45, 150)`

**2.** `prop.table(table(x), margin = 1)` calculates proportions:

    a) Across all cells (total = 1)    b) Within each row    c) Within each column    d) Divided by the number of columns

**3.** A Poisson distribution with lambda = 4 models an event that happens on average 4 times per interval. What is the relationship between the mean and variance of this distribution?

**4.** Write the R code to calculate the probability of observing **exactly 2** events from a Poisson distribution with lambda = 5.

**5.** True or False: The Poisson distribution is appropriate for modelling continuous measurements like protein concentration.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** b) `prop.test(45, 150)` — This is the correct function for a confidence interval on a proportion. `x` = number of successes, `n` = total observations.

**2.** b) Within each row — `margin = 1` calculates proportions so that each **row** sums to 1. Use this to compare groups when rows represent different categories.

**3.** They are **equal** — in the Poisson distribution, the mean (λ) equals the variance. This is a defining property of the Poisson. If your data shows variance much larger than the mean, a negative binomial distribution may be more appropriate.

**4.** `dpois(2, lambda = 5)` — `dpois` gives the probability of exactly `k` events; here k = 2, lambda = 5.

**5.** **False** — The Poisson distribution models **count data** (discrete, non-negative integers): number of mutations, number of events per unit time, number of organisms in a quadrat. Continuous measurements like concentration are modelled with the normal distribution or others.
:::

------------------------------------------------------------------------

## Lab 10 Checklist

Before you leave, make sure you can:

- [ ] Create a frequency table with `table()` and convert to proportions with `prop.table()`
- [ ] Build a two-way contingency table and calculate row or column proportions
- [ ] Use `prop.test()` to calculate a 95% confidence interval for a proportion
- [ ] Test whether an observed proportion is consistent with a known expected value
- [ ] Explain what lambda represents in a Poisson distribution
- [ ] Use `dpois()` for exact probabilities and `ppois()` for cumulative probabilities
- [ ] Simulate Poisson data with `rpois()` and compare to theoretical expectations

::: callout-tip
## Bonus Challenge

The spontaneous mutation rate of *E. coli* is approximately 1 × 10⁻³ mutations per genome replication. If we observe 50,000 replication events, what lambda would we use? Calculate P(0 mutations), P(exactly 50 mutations), and P(more than 100 mutations). Plot the distribution.
:::

------------------------------------------------------------------------

## Before Next Class (Wednesday)

- Wednesday (Lab 11): Hypothesis Testing and the Null Hypothesis — we go deeper on p-values, Type I/II errors, and effect sizes
- Read **Wilke 2021 Ch. 1, 2** (link on Canvas)

::: callout-important
## Quiz 3 — This Friday!

Quiz 3 covers **Labs 10 and 11**: frequency tables, proportions, CI for proportions, hypothesis testing, and the Poisson distribution.

Best preparation: be able to explain what a p-value means in plain English and know when to use `prop.test()` vs `t.test()`.
:::

------------------------------------------------------------------------
