---
title: "Lab 11: Hypothesis Testing & the Null Hypothesis"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "September 16, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Does This Antibiotic Actually Work?

Every clinical drug trial, every laboratory experiment comparing treatment to control, every epidemiological study asking whether a risk factor matters — all of these rest on **hypothesis testing**. It is arguably the most important inferential tool in science.

In this lab we build a thorough understanding of the hypothesis testing framework: what the null hypothesis means, what a p-value actually tells us, the consequences of getting the decision wrong, and how to measure whether an effect is not just statistically significant but biologically meaningful.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. State null and alternative hypotheses precisely for a biological question
2. Explain a p-value in plain language without common misconceptions
3. Distinguish Type I and Type II errors and explain their consequences
4. Run one-sample and two-sample t-tests for biological data
5. Compute and interpret Cohen's d as a measure of effect size
6. Understand the difference between statistical significance and practical importance
:::

------------------------------------------------------------------------

## Part 1: The Hypothesis Testing Framework

### The Logic

Scientists use hypothesis testing to decide whether observed data are consistent with a default assumption (the **null hypothesis**) or whether the data provide enough evidence to conclude something different is happening.

The framework always follows this structure:

::: callout-important
## The Four Steps of Hypothesis Testing

**Step 1 — State the hypotheses**
H₀ (null hypothesis): the default assumption — usually "no effect" or "no difference"
H₁ (alternative hypothesis): what you are trying to find evidence for

**Step 2 — Choose a significance level**
Conventionally α = 0.05. This is the probability of rejecting H₀ when it is actually true that we are willing to accept.

**Step 3 — Compute the test statistic and p-value**
Use the appropriate R function for your data type.

**Step 4 — Make a decision**
If p < α: reject H₀ — the data are unlikely under the null hypothesis
If p ≥ α: fail to reject H₀ — insufficient evidence to conclude H₁
:::

### A Biological Example

A new antibiotic is claimed to reduce bacterial colony counts. In untreated cultures, the mean colony count is 500 CFU/mL (CFU = colony forming units). You treat 20 cultures with the antibiotic and measure colony counts.

```{r}
# Colony counts (CFU/mL) from 20 antibiotic-treated cultures
set.seed(101)
treated_cfu <- round(rnorm(20, mean = 460, sd = 55))
treated_cfu
```

```{r}
cat("Sample mean:", round(mean(treated_cfu), 1), "CFU/mL\n")
cat("Sample SD:  ", round(sd(treated_cfu), 1), "CFU/mL\n")
cat("n =", length(treated_cfu), "\n")
```

The mean in our treated sample is lower than 500 — but is this a real effect or just random variation?

**H₀**: The mean CFU/mL in treated cultures = 500 (no effect)
**H₁**: The mean CFU/mL in treated cultures < 500 (antibiotic reduces colonies)

```{r}
# One-sample t-test against mu = 500, one-tailed (less than)
t.test(treated_cfu, mu = 500, alternative = "less")
```

------------------------------------------------------------------------

## Part 2: Understanding the p-value

The p-value is one of the most misunderstood quantities in science. Let's be precise.

::: callout-important
## What a p-value IS and IS NOT

**A p-value IS:**
The probability of obtaining a test statistic as extreme as (or more extreme than) the one observed, **assuming the null hypothesis is true**.

**A p-value IS NOT:**
- The probability that H₀ is true
- The probability that your result occurred by chance
- A measure of the size or importance of an effect
- Proof of anything

A small p-value means: "This result would be very unusual if H₀ were true." That is all it means.
:::

### Visualising the p-value

```{r}
# Visualise what the p-value represents
# Under H₀: t-distribution with df = 19
x <- seq(-5, 5, length = 500)
plot(x, dt(x, df = 19),
     type = "l", lwd = 2,
     col  = "#2C5F8A",
     xlab = "t statistic",
     ylab = "Density",
     main = "t-Distribution Under H₀ (df = 19)\nShaded area = p-value")

# Calculate observed t
t_obs <- (mean(treated_cfu) - 500) / (sd(treated_cfu) / sqrt(20))
cat("Observed t-statistic:", round(t_obs, 3), "\n")

# Shade the p-value region (left tail)
x_shade <- seq(-5, t_obs, length = 200)
polygon(c(x_shade, rev(x_shade)),
        c(dt(x_shade, df = 19), rep(0, 200)),
        col = "#C8102E60", border = NA)
abline(v = t_obs, col = "#C8102E", lty = 2, lwd = 2)
text(t_obs - 0.3, 0.3, paste0("t = ", round(t_obs, 2)),
     col = "#C8102E", adj = 1)
```

------------------------------------------------------------------------

## Part 3: Type I and Type II Errors

Every decision in hypothesis testing can be wrong in one of two ways:

```{r, echo=FALSE}
cat("
                     TRUE STATE OF THE WORLD
                  H₀ is true      H₀ is false
              ┌──────────────────────────────────┐
 OUR   Reject │ TYPE I ERROR   │ Correct decision │
DECISION  H₀  │ (False positive) │  (True positive) │
              │ Probability = α  │ Power = 1 - β   │
              ├──────────────────┼──────────────────┤
       Fail   │ Correct decision │ TYPE II ERROR   │
       to     │  (True negative) │ (False negative) │
       reject │                  │ Probability = β  │
              └──────────────────────────────────┘
")
```

::: callout-warning
## The Consequences of Each Error Type

**Type I Error (false positive, probability = α):**
You conclude the antibiotic works when it does not. Consequence: a useless drug enters clinical use; patients are exposed to side effects with no benefit.

**Type II Error (false negative, probability = β):**
You conclude the antibiotic does not work when it actually does. Consequence: an effective treatment is abandoned; patients who could have been helped are not.

Reducing α makes Type I errors less likely but increases Type II errors. The only way to reduce both simultaneously is to **increase sample size**.
:::

### Statistical Power

**Power** (= 1 − β) is the probability of correctly detecting an effect when it truly exists. In biology, we conventionally aim for 80% power (β = 0.20).

```{r}
# The power.t.test() function calculates required sample size
# Suppose the true mean reduction is 50 CFU/mL (delta = 50)
# SD from our pilot data ~ 55 CFU/mL
# We want 80% power at alpha = 0.05 (one-sided)

power.t.test(delta   = 50,
             sd      = 55,
             sig.level  = 0.05,
             power   = 0.80,
             type    = "one.sample",
             alternative = "one.sided")
```

::: callout-note
## Reading `power.t.test()` Output

The output tells you the **minimum sample size** needed to detect an effect of the given size with the specified power. If our true effect is 50 CFU/mL reduction and SD is 55, we need approximately this many treated cultures to have an 80% chance of detecting it.

Planning sample sizes before data collection is called **a priori power analysis** and is expected in scientific publications.
:::

------------------------------------------------------------------------

## Part 4: Two-Sample t-test — Comparing Treatment and Control

More commonly, we compare two independent groups. A new enzyme inhibitor is tested against a vehicle control in a cell viability assay:

```{r}
# Cell viability (% of control) in two groups
set.seed(202)
control    <- round(rnorm(25, mean = 100, sd = 12), 1)
inhibitor  <- round(rnorm(25, mean = 85,  sd = 14), 1)

cat("Control group:   mean =", round(mean(control), 1),
    "  SD =", round(sd(control), 1), "\n")
cat("Inhibitor group: mean =", round(mean(inhibitor), 1),
    "  SD =", round(sd(inhibitor), 1), "\n")
```

```{r}
# H₀: mean viability is the same in both groups
# H₁: inhibitor reduces viability (mean inhibitor < mean control)
result <- t.test(inhibitor, control,
                 alternative = "less",
                 var.equal   = FALSE)   # Welch's t-test — does not assume equal variances
result
```

```{r}
# Visualise
boxplot(list(Control = control, Inhibitor = inhibitor),
        col  = c("#A8C4D9", "#C8102E60"),
        ylab = "Cell Viability (% of control)",
        main = "Effect of Enzyme Inhibitor on Cell Viability\n(n = 25 per group)")
stripchart(list(Control = control, Inhibitor = inhibitor),
           method   = "jitter",
           vertical = TRUE,
           add      = TRUE,
           pch      = 16, cex = 0.7,
           col      = c("#2C5F8A", "#8B0000"))
```

------------------------------------------------------------------------

## Part 5: Effect Size — Cohen's d

Statistical significance tells you whether an effect is real; **effect size** tells you how big it is. A study with 10,000 samples can find statistically significant differences that are biologically trivial.

**Cohen's d** measures the standardised difference between two means:

d = (mean₁ − mean₂) / pooled SD

```{r}
# Calculate Cohen's d manually
cohens_d <- function(x, y) {
  pooled_sd <- sqrt(((length(x) - 1) * var(x) + (length(y) - 1) * var(y)) /
                    (length(x) + length(y) - 2))
  d <- (mean(x) - mean(y)) / pooled_sd
  return(d)
}

d <- cohens_d(control, inhibitor)
cat("Cohen's d:", round(d, 3), "\n")
```

::: callout-note
## Interpreting Cohen's d

| d value | Conventional interpretation |
|---------|---------------------------|
| 0.2 | Small effect |
| 0.5 | Medium effect |
| 0.8 | Large effect |

These thresholds come from Cohen (1988) and are widely used in biology. A large Cohen's d combined with a small p-value gives you the strongest evidence that an effect is both real and meaningful.

Our inhibitor shows d = 1.08 — a large effect. Even if the p-value were borderline, this magnitude of difference is biologically relevant for a cell viability assay.
:::

------------------------------------------------------------------------

## Part 6: One-Sided vs Two-Sided Tests

```{r}
# Same data, two-sided test (H₁: means are different, direction unspecified)
t.test(inhibitor, control, alternative = "two.sided", var.equal = FALSE)
```

::: callout-tip
## When to Use Each

Use a **two-sided test** (default) unless you have a strong prior biological reason to predict the direction of the effect before seeing the data.

Pre-registering a one-sided test because you believe a drug will reduce (not increase) cell viability is scientifically defensible. Switching to a one-sided test after seeing that your effect went in a convenient direction is not — this is sometimes called "p-hacking."
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** A researcher tests whether a new fertiliser changes plant height. H₀: mean height = 30 cm. After treating 20 plants they get p = 0.08. What is the correct conclusion at α = 0.05?

    a) Reject H₀; the fertiliser works    b) Accept H₀; the fertiliser has no effect    c) Fail to reject H₀; insufficient evidence    d) The experiment was inconclusive and must be repeated

**2.** A drug trial concludes the drug is ineffective (fails to reject H₀), but the drug actually does work. What type of error is this?

    a) Type I error    b) Type II error    c) Sampling error    d) Measurement error

**3.** Cohen's d = 0.15 for a comparison between two treatment groups. How would you describe this effect?

**4.** You run `t.test(group_A, group_B, alternative = "two.sided")`. What hypothesis does this test?

**5.** True or False: A p-value of 0.001 means there is a 0.1% probability that the null hypothesis is true.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** c) Fail to reject H₀ — p = 0.08 is greater than α = 0.05, so we do not have sufficient evidence to reject the null hypothesis. We cannot "accept" H₀ — we simply lack evidence against it. We also cannot conclude "no effect" — we can only say the evidence is insufficient.

**2.** b) Type II error — A false negative: the effect exists but we failed to detect it. This is also called a "miss." Its probability is β, and 1 − β is the statistical power of the test.

**3.** A small effect — Cohen's d = 0.15 is below the conventional threshold of 0.2 for a small effect. Even if the p-value were < 0.05, this magnitude of difference may not be biologically meaningful.

**4.** H₀: the means of group_A and group_B are equal, versus H₁: the means are different (in either direction). The two-sided test does not specify which group should be larger.

**5.** **False** — This is one of the most common misinterpretations of the p-value. A p-value is the probability of the data (or more extreme data) given that H₀ is true — NOT the probability that H₀ is true. Determining the probability of a hypothesis requires a different framework (Bayesian statistics).
:::

------------------------------------------------------------------------

## Lab 11 Checklist

Before you leave, make sure you can:

- [ ] Write H₀ and H₁ for a biological question in precise statistical language
- [ ] Explain what a p-value means without any of the common misconceptions
- [ ] Distinguish between Type I (false positive) and Type II (false negative) errors
- [ ] Run a one-sample t-test with `t.test(x, mu = value)`
- [ ] Run a two-sample Welch's t-test with `t.test(x, y, var.equal = FALSE)`
- [ ] Calculate and interpret Cohen's d as an effect size
- [ ] Explain why statistical significance alone is insufficient to claim a result is important

::: callout-tip
## Bonus Challenge

Using `power.t.test()`, calculate how many samples per group you would need to detect a Cohen's d of 0.5 with 90% power at α = 0.05 (two-sided). How does this change if you only need 70% power?
:::

------------------------------------------------------------------------

## Before Next Class (Quiz Friday, then Week 6)

- **Friday: Quiz 3** — covers frequency data, proportions, Poisson, and hypothesis testing (Labs 10–11)
- After the quiz, **Week 6** begins data visualisation
- Read **Wilke 2021 Ch. 5 and 7** over the weekend

------------------------------------------------------------------------
