---
title: "Lab 17: Strip Charts, Violin Plots, Paired t-test & Two-Sample t-test"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "October 5, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Comparing Two Groups: Visualisation and Inference Together

In experimental biology, the most common analytical question is: **does the treatment group differ from the control?** This lab builds the complete workflow — first visualising the data properly, then applying the appropriate t-test, and finally interpreting the result in biological context.

We cover two visualisation methods (strip charts and violin plots) that show all the data, and two types of t-test depending on whether the groups are independent or paired.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Produce strip charts with `stripchart()` and ggplot2 `geom_jitter()`
2. Produce violin plots with ggplot2 `geom_violin()`
3. Determine when to use an independent vs paired t-test
4. Run a paired t-test with `t.test(x, y, paired = TRUE)`
5. Run a two-sample Welch's t-test with `t.test(x, y, var.equal = FALSE)`
6. Report results in the format expected in a scientific paper
:::

------------------------------------------------------------------------

## Part 1: Why Visualise Before Testing?

Statistical tests summarise data into a single number (the p-value). Visualising the raw data first prevents you from being misled by that summary.

```{r}
library(ggplot2)

# Three datasets with identical means, SDs — but very different shapes
set.seed(42)
n <- 30

data_A <- rnorm(n, mean = 10, sd = 2)
data_B <- c(rnorm(15, 8, 0.5), rnorm(15, 12, 0.5))   # Bimodal
data_C <- c(rep(10, 20), rnorm(10, 10, 6))              # Outlier-heavy

df_demo <- data.frame(
  value = c(data_A, data_B, data_C),
  group = rep(c("A", "B", "C"), each = n)
)

# All three look identical as means ± SD
tapply(df_demo$value, df_demo$group, function(x) round(c(mean = mean(x), sd = sd(x)), 2))
```

```{r}
# But look very different when you plot the raw data
ggplot(df_demo, aes(x = group, y = value, colour = group)) +
  geom_jitter(width = 0.15, size = 2, alpha = 0.8) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 4, colour = "black") +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2, colour = "black") +
  labs(title  = "Three Datasets with Identical Means and SDs",
       subtitle = "Black diamond = mean ± SE",
       x = "Dataset", y = "Value") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

::: callout-warning
## Always Plot Raw Data

Groups B and C have the same summary statistics as A, but:
- B is bimodal (two subpopulations — perhaps treated and untreated cells mixed by accident)
- C has heavy outliers that are dragging the mean

A t-test would give almost identical p-values for all three comparisons. Only the plot reveals these problems. **Plot first, test second.**
:::

------------------------------------------------------------------------

## Part 2: Strip Charts

A **strip chart** (also called a dot plot) shows every individual data point. It is ideal for small to moderate sample sizes (n < 100).

```{r}
set.seed(303)
# Drug efficacy experiment: bacterial growth inhibition (mm zone of inhibition)
control    <- round(rnorm(20, mean = 0,    sd = 1.5))   # No drug
antibiotic <- round(rnorm(20, mean = 12.4, sd = 2.8))   # Drug added

# Base R stripchart
stripchart(list(Control = control, Antibiotic = antibiotic),
           method   = "jitter",
           vertical = TRUE,
           pch      = 16,
           cex      = 1.2,
           col      = c("#2C5F8A", "#C8102E"),
           ylab     = "Zone of Inhibition (mm)",
           main     = "Antibiotic Efficacy — Strip Chart")

# Add means ± SE
means <- c(mean(control), mean(antibiotic))
ses   <- c(sd(control)/sqrt(20), sd(antibiotic)/sqrt(20))
points(1:2, means, pch = 18, cex = 2.5, col = "black")
arrows(1:2, means - ses, 1:2, means + ses,
       angle = 90, code = 3, length = 0.08, lwd = 2, col = "black")
```

### Strip Charts with ggplot2

```{r}
# Combine into a data frame for ggplot2
growth_df <- data.frame(
  inhibition = c(control, antibiotic),
  group      = rep(c("Control", "Antibiotic"), each = 20)
)

ggplot(growth_df, aes(x = group, y = inhibition, colour = group)) +
  geom_jitter(width = 0.12, size = 2.5, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar",
               width = 0.3, fatten = 2, colour = "black") +
  scale_colour_manual(values = c("Control" = "#2C5F8A", "Antibiotic" = "#C8102E")) +
  labs(title = "Zone of Inhibition by Treatment",
       x     = "Treatment",
       y     = "Zone of Inhibition (mm)") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

------------------------------------------------------------------------

## Part 3: Violin Plots

A **violin plot** shows the full probability distribution of the data — essentially a smoothed density estimate mirrored on both sides. It is more informative than a box plot for revealing the shape of a distribution.

```{r}
set.seed(55)
# Cell viability (% of control) after three drug treatments
cell_df <- data.frame(
  viability = c(rnorm(30, 100, 8),     # Vehicle control
                rnorm(30,  72, 12),    # Drug X (1 uM)
                rnorm(30,  45, 15)),   # Drug X (10 uM)
  treatment = rep(c("Vehicle", "Drug 1uM", "Drug 10uM"), each = 30)
)

# Set factor order for logical display
cell_df$treatment <- factor(cell_df$treatment,
                             levels = c("Vehicle", "Drug 1uM", "Drug 10uM"))

ggplot(cell_df, aes(x = treatment, y = viability, fill = treatment)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  geom_jitter(width = 0.05, size = 1.2, alpha = 0.5, colour = "grey30") +
  scale_fill_manual(values = c("Vehicle"    = "#2C5F8A",
                               "Drug 1uM"   = "#5B8DB8",
                               "Drug 10uM"  = "#C8102E")) +
  labs(title = "Cell Viability Following Drug X Treatment",
       x     = "Treatment",
       y     = "Cell Viability (% of control)") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

::: callout-tip
## Violin + Box Plot Combination

Layering a narrow box plot inside a violin plot gives you:
- The **distribution shape** from the violin
- The **median, IQR, and outlier information** from the box plot
- The **individual data points** from `geom_jitter()`

This combination is increasingly expected in high-impact biology journals.
:::

------------------------------------------------------------------------

## Part 4: The Independent Two-Sample t-test

Use this test when comparing two **independent** groups — different subjects, cells, or experimental units in each group.

```{r}
# Return to the antibiotic example
# H₀: mean zone of inhibition is the same in control and antibiotic groups
# H₁: antibiotic produces a larger zone of inhibition

result_indep <- t.test(antibiotic, control,
                       alternative = "greater",   # We predict antibiotic > control
                       var.equal   = FALSE)        # Welch's t-test — safer default
result_indep
```

::: callout-note
## Reporting t-test Results

The conventional scientific format for reporting a t-test result is:

**"The antibiotic produced significantly larger zones of inhibition than the control (t(df) = t-value, p = p-value)."**

For example: *"The antibiotic produced significantly larger zones of inhibition than the control (t(29.4) = 21.3, p < 0.001)."*

Always include: the test statistic (t), degrees of freedom (df), and p-value. If effect size was calculated (Cohen's d), include that too.
:::

------------------------------------------------------------------------

## Part 5: The Paired t-test

Use a paired t-test when each observation in group 1 is **matched** to a specific observation in group 2. This is common in:

- Before/after designs (same subject measured twice)
- Matched case-control studies
- Left/right comparisons on the same organism
- Replicate wells on the same plate

Pairing removes between-subject variability, making the test more powerful.

```{r}
# Before/after experiment: blood glucose (mmol/L) in 18 patients
# measured before and 2 hours after an oral glucose tolerance test (OGTT)
set.seed(888)
n_patients <- 18
glucose_fasting  <- round(rnorm(n_patients, mean = 5.2, sd = 0.5), 1)
glucose_2hr      <- glucose_fasting + round(rnorm(n_patients, mean = 3.8, sd = 1.2), 1)

cat("Fasting glucose:   mean =", round(mean(glucose_fasting), 2), "mmol/L\n")
cat("2-hour glucose:    mean =", round(mean(glucose_2hr),     2), "mmol/L\n")
cat("Mean difference:         ", round(mean(glucose_2hr - glucose_fasting), 2), "mmol/L\n")
```

```{r}
# Paired t-test
# H₀: mean glucose does not change after OGTT
# H₁: glucose increases after OGTT
result_paired <- t.test(glucose_2hr, glucose_fasting,
                        paired      = TRUE,
                        alternative = "greater")
result_paired
```

```{r}
# Visualise the paired nature of the data
patient_df <- data.frame(
  glucose  = c(glucose_fasting, glucose_2hr),
  timepoint = rep(c("Fasting", "2-Hour Post-OGTT"), each = n_patients),
  patient  = rep(1:n_patients, 2)
)

patient_df$timepoint <- factor(patient_df$timepoint,
                                levels = c("Fasting", "2-Hour Post-OGTT"))

ggplot(patient_df, aes(x = timepoint, y = glucose)) +
  geom_line(aes(group = patient), colour = "grey70", linewidth = 0.5) +
  geom_point(aes(colour = timepoint), size = 3) +
  scale_colour_manual(values = c("Fasting"          = "#2C5F8A",
                                 "2-Hour Post-OGTT" = "#C8102E")) +
  stat_summary(fun = mean, geom = "point", shape = 18,
               size = 5, colour = "black") +
  labs(title    = "Blood Glucose: Oral Glucose Tolerance Test",
       subtitle = paste0("n = ", n_patients, " patients; paired t-test, p = ",
                         formatC(result_paired$p.value, format = "e", digits = 2)),
       x        = "Timepoint",
       y        = "Blood Glucose (mmol/L)") +
  theme_classic(base_size = 13) +
  theme(legend.position = "none")
```

::: callout-important
## Paired vs Independent: Which Do You Use?

| Design | Test | Why |
|--------|------|-----|
| Different subjects in each group | Independent (Welch's) t-test | Groups share no observations |
| Same subjects measured twice | Paired t-test | Removes within-subject variability |
| Matched controls | Paired t-test | Each case has a specific matched control |

Using a paired test when data are independent (or vice versa) gives the wrong answer. **Always check your experimental design before choosing a test.**
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** You are comparing body weight in 25 mice before and after a 4-week dietary intervention (the same 25 mice measured twice). Which test is appropriate?

    a) Independent two-sample t-test    b) Paired t-test    c) One-sample t-test    d) Chi-square test

**2.** What does a violin plot show that a box plot does not?

**3.** In a paired t-test, the test is effectively performed on:

    a) The raw values in each group    b) The differences between paired observations    c) The ranks of the observations    d) The log-transformed values

**4.** You run `t.test(treated, control, var.equal = FALSE)`. What does `var.equal = FALSE` specify?

**5.** True or False: A strip chart (dot plot) is more informative than a mean ± SD bar chart for small sample sizes (n < 20).
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** b) Paired t-test — The same 25 mice are measured twice: once before and once after the intervention. Each mouse's two measurements are paired. The paired t-test removes between-mouse variability (e.g., natural size differences) and focuses on the change within each mouse.

**2.** A violin plot shows the **full shape of the distribution** — including whether it is unimodal, bimodal, symmetric, or skewed. A box plot only shows the median, quartiles, and outliers. A bimodal distribution (two peaks) can hide completely inside a box plot.

**3.** b) The differences between paired observations — a paired t-test computes d = x₁ − x₂ for each pair and then runs a one-sample t-test asking whether the mean difference is significantly different from zero. This is mathematically equivalent to `t.test(x1 - x2, mu = 0)`.

**4.** `var.equal = FALSE` specifies **Welch's t-test**, which does not assume equal variances between the two groups. This is the safer default because violation of equal variances inflates Type I error in the standard Student's t-test. Use `var.equal = TRUE` only if you have strong reason to believe the variances are equal (which you should verify with Levene's test — Lab 18).

**5.** **True** — For small samples, individual data points are visible and meaningful. A mean ± SD bar chart hides the distribution entirely — with n = 5, for example, you cannot tell if the data are normally distributed, skewed, or contain outliers. Strip charts reveal these features directly.
:::

------------------------------------------------------------------------

## Lab 17 Checklist

Before you leave, make sure you can:

- [ ] Explain why you should plot data before running a statistical test
- [ ] Produce a strip chart with `stripchart()` and with `geom_jitter()`
- [ ] Produce a violin plot with `geom_violin()` and overlay a box plot and jittered points
- [ ] Determine whether a paired or independent t-test is appropriate for a given design
- [ ] Run `t.test(x, y, paired = TRUE)` and interpret the output
- [ ] Run `t.test(x, y, var.equal = FALSE)` (Welch's) and interpret the output
- [ ] Report t-test results in the standard scientific format: t(df) = value, p = value

::: callout-tip
## Bonus Challenge

Design your own experiment: choose a biological question involving two groups (e.g., antibiotic A vs antibiotic B, treated vs untreated cells, gene expression before vs after induction). Simulate realistic data using `rnorm()` with biologically plausible means and SDs. Produce a publication-ready violin + jitter plot in ggplot2, run the appropriate t-test, and write a one-sentence conclusion in scientific language.
:::

------------------------------------------------------------------------

## Before Next Class (Wednesday)

- Wednesday (Lab 18): Confidence intervals from t-tests, and Levene's test for equal variances
- Read **R for Data Science Ch. 9–10**

------------------------------------------------------------------------
