---
title: "Lab 22: Simple Linear Regression"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "October 19, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## From Correlation to Prediction

Correlation tells you that two variables are related. **Regression** goes further: it describes the relationship with a mathematical equation, allowing you to predict one variable from another and to test whether the predictor explains a significant portion of the variation.

Simple linear regression fits a straight line: **y = β₀ + β₁x + ε**

- β₀ = intercept (value of y when x = 0)
- β₁ = slope (change in y for a one-unit increase in x)
- ε = residual error (what the model cannot explain)

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Fit a simple linear regression with `lm()` in R
2. Interpret the slope, intercept, R², and F-test from `summary(lm())`
3. Check regression assumptions with diagnostic plots
4. Plot a regression line with 95% confidence and prediction bands
5. Make predictions from a fitted model with `predict()`
6. Describe the difference between a confidence interval and a prediction interval
:::

------------------------------------------------------------------------

## Part 1: The Bradford Protein Assay — A Classic Biological Calibration Curve

The Bradford assay estimates protein concentration by measuring absorbance at 595 nm. A standard curve is produced by measuring known concentrations of BSA (bovine serum albumin) — this is a textbook linear regression problem in biochemistry.

```{r}
# BSA standard concentrations (ug/mL) and measured absorbance (A595)
bsa_conc <- c(0, 50, 100, 150, 200, 250, 300, 350, 400)
absorbance <- c(0.041, 0.189, 0.352, 0.508, 0.667, 0.821, 0.979, 1.128, 1.276) +
              rnorm(9, 0, 0.018)

# Always visualise first
plot(bsa_conc, absorbance,
     pch  = 16,
     col  = "#2C5F8A",
     cex  = 1.4,
     xlab = "BSA Concentration (ug/mL)",
     ylab = "Absorbance at 595 nm",
     main = "Bradford Assay Standard Curve")
```

```{r}
# Fit the regression model
bradford_model <- lm(absorbance ~ bsa_conc)

# Examine the full summary
summary(bradford_model)
```

::: callout-important
## Reading the `summary(lm())` Output

**Coefficients table:**

| Term | Estimate | Meaning |
|------|----------|---------|
| `(Intercept)` | β₀ | Expected absorbance when [BSA] = 0 (should be near 0 for blank) |
| `bsa_conc` | β₁ | Increase in absorbance per 1 ug/mL increase in BSA |

**Overall model fit:**

| Statistic | Meaning |
|-----------|---------|
| `Multiple R-squared` | Proportion of variation in y explained by x |
| `F-statistic, p-value` | Overall test: does x significantly predict y? (H₀: β₁ = 0) |
| `Residual standard error` | Average distance of data points from the fitted line |

An R² of 0.998 means 99.8% of the variation in absorbance is explained by BSA concentration — ideal for a calibration curve.
:::

------------------------------------------------------------------------

## Part 2: Plotting the Regression Line with Confidence Bands

```{r}
library(ggplot2)

# Create a data frame for ggplot
bradford_df <- data.frame(conc = bsa_conc, abs = absorbance)

# New x values for smooth confidence band
new_conc <- data.frame(bsa_conc = seq(0, 400, by = 5))

# Confidence band (uncertainty in the mean prediction)
ci_band <- predict(bradford_model, newdata = new_conc,
                   interval = "confidence", level = 0.95)

# Prediction band (uncertainty for a new single measurement)
pi_band <- predict(bradford_model, newdata = new_conc,
                   interval = "prediction", level = 0.95)

band_df <- data.frame(new_conc,
                       fit    = ci_band[, "fit"],
                       ci_lo  = ci_band[, "lwr"],
                       ci_hi  = ci_band[, "upr"],
                       pi_lo  = pi_band[, "lwr"],
                       pi_hi  = pi_band[, "upr"])

ggplot(bradford_df, aes(x = conc, y = abs)) +
  geom_ribbon(data = band_df, aes(x = bsa_conc, ymin = pi_lo, ymax = pi_hi),
              fill = "#A8C4D9", alpha = 0.4, inherit.aes = FALSE) +
  geom_ribbon(data = band_df, aes(x = bsa_conc, ymin = ci_lo, ymax = ci_hi),
              fill = "#5B8DB8", alpha = 0.5, inherit.aes = FALSE) +
  geom_line(data = band_df, aes(x = bsa_conc, y = fit),
            colour = "#2C5F8A", linewidth = 1.2, inherit.aes = FALSE) +
  geom_point(colour = "#2C5F8A", size = 3) +
  annotate("text", x = 50, y = 1.2,
           label = paste0("R² = ", round(summary(bradford_model)$r.squared, 4)),
           hjust = 0, size = 4) +
  labs(title    = "Bradford Assay Standard Curve",
       subtitle = "Dark band = 95% CI for mean; light band = 95% PI for new observation",
       x        = "BSA Concentration (ug/mL)",
       y        = "Absorbance (595 nm)") +
  theme_classic(base_size = 13)
```

::: callout-note
## Confidence Interval vs Prediction Interval

**Confidence interval (CI)**: the plausible range for the **mean** absorbance at a given concentration — gets narrower with more data.

**Prediction interval (PI)**: the plausible range for the absorbance of a **single new observation** at that concentration — always wider than the CI because it incorporates measurement-to-measurement variability.

When using a standard curve to estimate an unknown protein concentration, the prediction interval is more honest — you are predicting a single new measurement, not the average of many.
:::

------------------------------------------------------------------------

## Part 3: Making Predictions

```{r}
# Three unknown samples were measured at these absorbances:
unknown_abs <- c(0.55, 0.88, 1.10)

# Invert the regression: concentration = (absorbance - intercept) / slope
intercept <- coef(bradford_model)[1]
slope     <- coef(bradford_model)[2]

predicted_conc <- (unknown_abs - intercept) / slope
cat("Predicted concentrations (ug/mL):\n")
print(round(predicted_conc, 1))
```

```{r}
# Or use predict() directly (predict absorbance from concentration)
# For inverse prediction, you can also use the equation above
# Let's confirm: what absorbance do we expect at 175 ug/mL?
predict(bradford_model,
        newdata  = data.frame(bsa_conc = 175),
        interval = "prediction")
```

------------------------------------------------------------------------

## Part 4: Checking Regression Assumptions

Linear regression assumes:

1. **Linearity** — the true relationship is linear
2. **Independence** — observations are independent
3. **Homoscedasticity** — residuals have constant variance
4. **Normality of residuals** — residuals are approximately normally distributed

```{r}
par(mfrow = c(2, 2))
plot(bradford_model)
par(mfrow = c(1, 1))
```

::: callout-tip
## Diagnosing the Four Plots

| Plot | What to look for |
|------|-----------------|
| **Residuals vs Fitted** | Points scattered randomly around 0 — no pattern. Curved pattern = nonlinearity. Funnel = heteroscedasticity. |
| **Normal QQ of residuals** | Points on the diagonal line — confirms normality of residuals. |
| **Scale-Location** | Square root of standardised residuals vs fitted — flat line = constant variance. |
| **Residuals vs Leverage** | Identifies influential points. Points outside Cook's distance lines have disproportionate influence on the slope. |
:::

------------------------------------------------------------------------

## Part 5: A Second Example — Allometric Scaling

Allometric scaling describes how physiological variables scale with body mass. Basal metabolic rate (BMR) scales approximately with body mass raised to the power 0.75 — on a log-log scale, this becomes a straight line.

```{r}
set.seed(88)
# Mammal body mass (kg) and BMR (kcal/day) — realistic scale range
body_mass <- c(0.02, 0.05, 0.1, 0.5, 1, 3, 10, 20, 50, 100, 200, 500, 1000, 3000, 5000)
bmr       <- 70 * (body_mass^0.75) * exp(rnorm(15, 0, 0.15))

# Linear regression on log-log scale
log_mass <- log10(body_mass)
log_bmr  <- log10(bmr)

allometry_model <- lm(log_bmr ~ log_mass)
summary(allometry_model)
```

```{r}
# The slope on log-log scale = the allometric exponent
cat("Allometric exponent (slope on log-log scale):",
    round(coef(allometry_model)[2], 3), "\n")
cat("Expected from Kleiber's law: ~0.75\n")
```

```{r}
plot(log_mass, log_bmr,
     pch  = 16, col = "#2C5F8A", cex = 1.4,
     xlab = "log₁₀ Body Mass (kg)",
     ylab = "log₁₀ BMR (kcal/day)",
     main = "Allometric Scaling of Basal Metabolic Rate")
abline(allometry_model, col = "#C8102E", lwd = 2)
legend("topleft",
       legend = c(paste0("slope = ", round(coef(allometry_model)[2], 3)),
                  paste0("R² = ", round(summary(allometry_model)$r.squared, 3))),
       bty = "n")
```

::: callout-note
## Log-Log Transformation in Allometry

Many biological relationships follow **power laws**: y = a × x^b. On a log-log scale, this becomes log(y) = log(a) + b × log(x) — a straight line. The slope b is the allometric exponent, and the intercept log(a) gives the proportionality constant.

This is why biologists often plot mass-related data on log-log axes — it linearises power-law relationships and makes regression analysis valid.
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** In the output of `summary(lm())`, the coefficient for the predictor has estimate = 0.0032 and p-value = 0.0001. What does this tell you?

**2.** A regression model gives R² = 0.45. Which statement is correct?

    a) 45% of observations are correctly predicted    b) The model explains 45% of the variance in y    c) The slope is 0.45    d) 45% of data points lie on the regression line

**3.** You run a Bradford assay and get an absorbance of 0.72 for an unknown sample. Explain step by step how you would use a linear regression model to estimate the protein concentration.

**4.** In a Residuals vs Fitted plot, the points form a U-shaped curve. What does this indicate and what should you do?

**5.** True or False: A prediction interval for a new observation is always narrower than the confidence interval for the mean at the same x value.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** The slope is 0.0032, meaning that for each one-unit increase in x, y increases by 0.0032 on average. The p-value of 0.0001 means this slope is highly significantly different from zero — we reject H₀: β₁ = 0. The predictor x significantly explains variation in y. However, the biological importance depends on the units and context, not just statistical significance.

**2.** b) The model explains 45% of the variance in y — R² (the coefficient of determination) is the proportion of variance in the response variable accounted for by the predictor. It does not mean 45% of predictions are "correct" (which has no clear meaning in regression) or that 45% of points lie on the line.

**3.** Step 1: Identify the model intercept (β₀) and slope (β₁) from `coef(model)`. Step 2: Set up the equation: absorbance = β₀ + β₁ × concentration. Step 3: Solve for concentration: concentration = (absorbance − β₀) / β₁ = (0.72 − β₀) / β₁. Step 4: Check that 0.72 falls within the range of your standard curve (do not extrapolate beyond it).

**4.** A U-shaped (curved) pattern in Residuals vs Fitted indicates **non-linearity** — the true relationship between x and y is not linear, even though you fitted a straight line. You should consider: (a) adding a quadratic term (x²) to the model, (b) log-transforming x or y, or (c) fitting a curve instead of a line. A linear model is not appropriate for these data.

**5.** False — The prediction interval is always **wider** than the confidence interval. The CI quantifies uncertainty in the mean prediction; the PI additionally accounts for individual observation variability (the residual variance). At the same x value, PI = CI width plus the residual standard error contribution. A PI will always be broader.
:::

------------------------------------------------------------------------

## Lab 22 Checklist

Before you leave, make sure you can:

- [ ] Fit a linear regression with `lm(y ~ x)` and view the summary
- [ ] Interpret the slope, intercept, R², and overall F-test p-value
- [ ] Use `predict()` with `interval = "confidence"` and `interval = "prediction"`
- [ ] Explain the difference between a CI and a PI for regression
- [ ] Produce a scatter plot with a regression line and confidence/prediction bands in ggplot2
- [ ] Interpret the four standard diagnostic plots (`plot(model)`)
- [ ] Apply log-log regression for power-law relationships and interpret the slope

::: callout-tip
## Bonus Challenge

The R built-in dataset `trees` contains girth, height, and volume for 31 black cherry trees. Fit a linear regression of `Volume` on `Girth`. Check all four diagnostic plots. Then try regressing log(Volume) on log(Girth) and compare R² and the diagnostic plots. Which model fits better? Write a two-sentence biological interpretation of the slope in each model.
:::

------------------------------------------------------------------------

## Before Next Class (Wednesday)

- Wednesday (Lab 23): Contingency tests — chi-square goodness of fit and tests of independence for genetic and epidemiological data
- Friday: **Quiz 6** — covers Labs 21–24 (correlation, regression, contingency tests, Hardy-Weinberg)

------------------------------------------------------------------------
