---
title: "Lab 9: Data Wrangling with dplyr"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "September 11, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Wrangling NFL Data with dplyr

Every real dataset needs to be cleaned, reorganized, and summarized before you can analyze it. **dplyr** (pronounced "dee-ply-er") is the most popular R package for this and once you learn its six main verbs, you'll use them in virtually every analysis you ever do.

Today we'll learn dplyr using NFL player stats, which come in exactly the kind of messy, multi-column format that real biological data does too.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Install and load packages using `install.packages()` and `library()`
2. Use the pipe operator `|>` to chain operations
3. Filter rows with `filter()`
4. Select columns with `select()`
5. Create new columns with `mutate()`
6. Sort data with `arrange()`
7. Summarize data with `summarize()` and `group_by()`
:::

------------------------------------------------------------------------

## Part 1: Installing and Loading Packages

R comes with a lot built in, but much of its power comes from **packages** — collections of extra functions written by the R community.

::: callout-important
## Installing vs. Loading a Package

**Install** = download to your computer (do this ONCE):
```r
install.packages("dplyr")
```

**Load** = activate for use in this session (do this EVERY time):
```r
library(dplyr)
```

Think of installing like buying a book, and loading like opening it to read.
:::

```{r}
# If you haven't installed dplyr yet, uncomment the line below:
# install.packages("dplyr")

library(dplyr)
```

------------------------------------------------------------------------

## Part 2: Our Dataset — 2025 NFL Season Stats

```{r}
# NFL skill player stats for the 2025-26 season (simulated)
nfl <- data.frame(
  player      = c("Patrick Mahomes", "Lamar Jackson", "Jalen Hurts",
                   "Josh Allen", "Justin Herbert", "Dak Prescott",
                   "Brock Purdy", "Tua Tagovailoa", "CJ Stroud", "Joe Burrow",
                   "Tyreek Hill", "Justin Jefferson", "CeeDee Lamb",
                   "Stefon Diggs", "Davante Adams", "DK Metcalf",
                   "Cooper Kupp", "Ja'Marr Chase", "Travis Kelce",
                   "Mark Andrews", "Darren Waller", "Sam LaPorta",
                   "Derrick Henry", "Christian McCaffrey", "Josh Jacobs",
                   "Tony Pollard", "Saquon Barkley", "Breece Hall",
                   "Isiah Pacheco", "Jonathan Taylor"),
  position    = c(rep("QB", 10), rep("WR", 8), rep("TE", 4), rep("RB", 8)),
  team        = c("KC", "BAL", "PHI", "BUF", "LAC", "DAL", "SF", "MIA", "HOU", "CIN",
                   "MIA", "MIN", "DAL", "BUF", "GB", "SEA", "LAR", "CIN",
                   "KC", "BAL", "LV", "DET",
                   "TEN", "SF", "GB", "DAL", "PHI", "NYJ", "KC", "IND"),
  conference  = c(rep("AFC", 1), rep("AFC", 1), rep("NFC", 1), rep("AFC", 1),
                   rep("AFC", 1), rep("NFC", 1), rep("NFC", 1), rep("AFC", 1),
                   rep("AFC", 1), rep("AFC", 1),
                   rep("AFC", 1), rep("NFC", 1), rep("NFC", 1), rep("AFC", 1),
                   rep("NFC", 1), rep("NFC", 1), rep("NFC", 1), rep("AFC", 1),
                   rep("AFC", 1), rep("AFC", 1), rep("AFC", 1), rep("NFC", 1),
                   rep("AFC", 1), rep("NFC", 1), rep("NFC", 1), rep("NFC", 1),
                   rep("NFC", 1), rep("AFC", 1), rep("AFC", 1), rep("AFC", 1)),
  games       = c(16, 17, 15, 17, 16, 14, 17, 15, 17, 13,
                   17, 17, 17, 14, 16, 17, 15, 17,
                   17, 16, 12, 17,
                   17, 14, 17, 17, 17, 16, 17, 11),
  yards       = c(4839, 3678, 3858, 4306, 3764, 3247, 4280, 3561, 4108, 2987,
                   1481, 1529, 1749, 1095, 1144, 1249, 1002, 1455,
                   984, 847, 762, 1130,
                   1458, 1459, 1166, 1024, 1499, 1187, 872, 897),
  touchdowns  = c(38, 24, 31, 40, 26, 22, 28, 24, 33, 19,
                   13, 10, 17, 9, 12, 11, 8, 14,
                   8, 9, 6, 10,
                   12, 11, 8, 7, 13, 9, 6, 7),
  pro_bowl    = c(TRUE, TRUE, TRUE, TRUE, FALSE, TRUE, FALSE, FALSE, TRUE, TRUE,
                   TRUE, TRUE, TRUE, FALSE, FALSE, FALSE, FALSE, TRUE,
                   TRUE, TRUE, FALSE, FALSE,
                   FALSE, TRUE, FALSE, FALSE, TRUE, FALSE, FALSE, FALSE)
)
```

Let's get a quick look at what we're working with:

```{r}
dim(nfl)
head(nfl)
```

------------------------------------------------------------------------

## Part 3: The Pipe Operator `|>`

Before learning the dplyr verbs, we need to understand the **pipe** `|>`. It takes the output of one function and passes it as the **first argument** to the next function.

```{r}
# Without pipe (hard to read, right?)
head(filter(nfl, position == "QB"), 3)
```

```{r}
# With pipe (reads left-to-right like a sentence)
nfl |> filter(position == "QB") |> head(3)
```

::: callout-tip
## Reading the Pipe as "then"
`nfl |> filter(position == "QB") |> head(3)`

Read this as: **"Take nfl, THEN filter to QBs only, THEN show the first 3 rows."**

The pipe makes your code read like a recipe, and you'll use it in almost every data analysis going forward.

**Keyboard shortcut**: Ctrl+Shift+M (Windows) / Cmd+Shift+M (MacOS)
:::

------------------------------------------------------------------------

## Part 4: `filter()` — Selecting Rows

`filter()` keeps only the rows that match your conditions.

```{r}
# All Pro Bowl players
nfl |> filter(pro_bowl == TRUE)
```

```{r}
# All receivers (WR and TE) with more than 1,000 yards
nfl |> filter(position %in% c("WR", "TE"), yards > 1000)
```

```{r}
# AFC players with 10 or more touchdowns
nfl |> filter(conference == "AFC", touchdowns >= 10)
```

::: callout-note
## Useful `filter()` Operators

| Operator | Meaning | Example |
|----------|---------|---------|
| `==` | Equal to | `position == "QB"` |
| `!=` | Not equal to | `team != "KC"` |
| `>`, `<`, `>=`, `<=` | Comparisons | `yards > 1000` |
| `&` | AND (both must be true) | `yards > 1000 & touchdowns > 10` |
| `\|` | OR (at least one true) | `position == "WR" \| position == "TE"` |
| `%in%` | Matches any in a list | `team %in% c("KC", "PHI", "SF")` |
| `is.na()` | Is missing | `is.na(yards)` |
:::

------------------------------------------------------------------------

## Part 5: `select()` — Choosing Columns

`select()` lets you keep only the columns you need:

```{r}
# Just player name, position, and touchdowns
nfl |> select(player, position, touchdowns)
```

```{r}
# Remove a column using -
nfl |> select(-pro_bowl) |> head(4)
```

```{r}
# Select a range of columns
nfl |> select(player:team) |> head(4)
```

```{r}
# Reorder columns and drop the rest
nfl |> select(player, team, position, yards, touchdowns) |> head(4)
```

------------------------------------------------------------------------

## Part 6: `mutate()` — Creating New Columns

`mutate()` adds new columns calculated from existing ones:

```{r}
# Yards per game
nfl |>
  mutate(yards_per_game = round(yards / games, 1)) |>
  select(player, position, games, yards, yards_per_game) |>
  head(8)
```

```{r}
# Touchdowns per game AND a performance tier
nfl |>
  mutate(
    td_per_game    = round(touchdowns / games, 2),
    performance    = case_when(
      touchdowns >= 30 ~ "Elite",
      touchdowns >= 15 ~ "Pro-Bowl Level",
      touchdowns >= 8  ~ "Solid Starter",
      TRUE             ~ "Developing"
    )
  ) |>
  select(player, position, touchdowns, td_per_game, performance) |>
  arrange(desc(touchdowns))
```

::: callout-tip
## `case_when()` — The Multi-Condition `ifelse()`
`case_when()` is like a series of if-else statements. Each line is:
`condition ~ value_to_assign`

The last line (`TRUE ~ ...`) is the "else" — catches everything that didn't match.
:::

------------------------------------------------------------------------

## Part 7: `arrange()` — Sorting

```{r}
# Sort by touchdowns, highest first
nfl |>
  select(player, position, touchdowns) |>
  arrange(desc(touchdowns))
```

```{r}
# Sort alphabetically by team, then by yards within each team
nfl |>
  select(player, team, yards) |>
  arrange(team, desc(yards)) |>
  head(10)
```

------------------------------------------------------------------------

## Part 8: `summarize()` and `group_by()` — The Power Combo

This is the most important dplyr pattern. `group_by()` splits the data into groups, and `summarize()` collapses each group into summary statistics.

```{r}
# Average yards and touchdowns by position
nfl |>
  group_by(position) |>
  summarize(
    n_players     = n(),
    avg_yards     = round(mean(yards), 0),
    avg_tds       = round(mean(touchdowns), 1),
    total_tds     = sum(touchdowns),
    pro_bowl_pct  = round(mean(pro_bowl) * 100, 1)
  )
```

```{r}
# Compare AFC vs NFC
nfl |>
  group_by(conference) |>
  summarize(
    players      = n(),
    avg_yards    = round(mean(yards), 0),
    avg_tds      = round(mean(touchdowns), 1),
    pro_bowlers  = sum(pro_bowl)
  )
```

```{r}
# Top team by total touchdowns
nfl |>
  group_by(team) |>
  summarize(team_tds = sum(touchdowns)) |>
  arrange(desc(team_tds)) |>
  head(8)
```

::: callout-note
## Useful `summarize()` Functions

| Function | What it does |
|----------|-------------|
| `n()` | Count rows in group |
| `mean(x)` | Average |
| `sum(x)` | Total |
| `median(x)` | Middle value |
| `sd(x)` | Standard deviation |
| `min(x)`, `max(x)` | Smallest / largest |
| `n_distinct(x)` | Count unique values |
:::

------------------------------------------------------------------------

## Part 9: Chaining Everything Together

The real power of dplyr is chaining multiple verbs into a readable pipeline:

```{r}
# Full analysis: Pro Bowl skill players in the AFC with 15+ TDs
# Show their yards per game, sorted by performance

nfl |>
  filter(conference == "AFC",
         pro_bowl == TRUE,
         touchdowns >= 10) |>
  mutate(ypg = round(yards / games, 1)) |>
  select(player, team, position, touchdowns, ypg) |>
  arrange(desc(touchdowns))
```

```{r}
# Summary: Average stats for Pro Bowl vs Non-Pro Bowl players
nfl |>
  group_by(pro_bowl) |>
  summarize(
    count     = n(),
    avg_yards = round(mean(yards)),
    avg_tds   = round(mean(touchdowns), 1),
    avg_games = round(mean(games), 1)
  )
```

```{r}
# Visualization: Touchdowns by position group
pos_summary <- nfl |>
  group_by(position) |>
  summarize(avg_tds = mean(touchdowns))

barplot(
  pos_summary$avg_tds,
  names.arg = pos_summary$position,
  col       = c("#013369", "#D50A0A", "#1B912F", "#F0D60F"),
  ylab      = "Average Touchdowns",
  main      = "Average Touchdowns by Position\n2025–26 NFL Season",
  border    = "white"
)
```

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together.*

::: callout-caution
## Knowledge Check Questions

**1.** What does the pipe operator `|>` do?

    a) Divides two numbers  b) Passes the result of one function into the next c) Creates a new column d) Filters rows

**2.** Which dplyr verb would you use to keep only NFL players who scored more than 10 touchdowns?

    a) `select()` b) `mutate()` c) `filter()` d) `arrange()`

**3.** You want to add a new column `ypg` (yards per game) to your data frame. Which verb do you use, and write the code?

**4.** What is the correct order for calculating average touchdowns per position?

  a) `summarize()` then `group_by()`
  b) `group_by()` then `summarize()`
  c) `filter()` then `group_by()`
  d) Order doesn't matter

**5.** Inside `summarize()`, which function counts the number of rows in each group?

  a) `count()`
  b) `length()`
  c) `n()`
  d) `tally()`

:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** b) Passes the result of one function into the next — The pipe `|>` takes whatever is on its left and feeds it as the first argument to the function on its right. It lets you chain steps together in a readable left-to-right sequence.

**2.** c) `filter()` — `filter()` selects **rows** based on conditions. Example: `nfl |> filter(touchdowns > 10)`.

**3.** `mutate()` — `nfl |> mutate(ypg = round(yards / games, 1))`. `mutate()` **creates or modifies columns** using expressions based on existing columns.

**4.** b) `group_by()` then `summarize()` — You must tell R what groups to use **before** you calculate the summaries. `group_by(position) |> summarize(avg_td = mean(touchdowns))`.

**5.** c) `n()` — Inside `summarize()`, `n()` counts the number of rows in each group. Note: `length()` also works on a column but `n()` is the idiomatic dplyr approach.
:::

------------------------------------------------------------------------

## Lab 9 Checklist

Before you leave, make sure you can:

- [ ] Load a package using `library()`
- [ ] Use the pipe `|>` to chain operations
- [ ] Filter rows using `filter()` with one and multiple conditions
- [ ] Select specific columns using `select()`
- [ ] Create new columns using `mutate()`
- [ ] Sort data using `arrange()` (and `desc()` for descending)
- [ ] Calculate group summaries using `group_by()` + `summarize()`
- [ ] Chain multiple dplyr verbs together in a single pipeline

::: callout-tip
## Bonus Challenge
Using the `nfl` dataset, find the top 5 players in **yards per game** (you'll need to create this column with `mutate()`). Then filter to only those who played at least 15 games. Show their name, team, position, and yards per game. Which position dominates the top 5?
:::

------------------------------------------------------------------------

## Before Next Class (Monday, Week 5)

- **Monday Sep 14** (Lab 10): Frequency Data & Confidence Intervals for Proportions
- Read **Wilke 2021 Ch. 1, 2, 5, 7** (links on Canvas)
- dplyr will be used in almost every lab from now on — get comfortable with the 6 verbs!

::: callout-note
## The 6 dplyr Verbs — Quick Reference

| Verb | What it does | Key argument |
|------|-------------|--------------|
| `filter()` | Keep rows matching conditions | Logical expressions |
| `select()` | Keep/drop columns | Column names |
| `mutate()` | Add/change columns | `new_col = expression` |
| `arrange()` | Sort rows | Column name(s), `desc()` |
| `summarize()` | Collapse to summary stats | `stat = function(col)` |
| `group_by()` | Split into groups before summarizing | Column to group by |
:::

------------------------------------------------------------------------
