---
title: "Lab 25: Multiple Regression & Model Diagnostics"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "October 26, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Beyond One Predictor

In Lab 22 we fitted a line through data with a single predictor. In most biological systems, the response variable is influenced by multiple factors simultaneously. **Multiple regression** models this: it estimates the independent contribution of each predictor while controlling for the others.

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Fit a multiple linear regression with `lm(y ~ x1 + x2 + x3)`
2. Interpret partial regression coefficients (the meaning of "controlling for other predictors")
3. Understand adjusted R² and why it differs from R²
4. Compare competing models using AIC with `AIC()`
5. Detect and respond to multicollinearity with `cor()` and VIF
6. Check all regression assumptions with diagnostic plots
:::

------------------------------------------------------------------------

## Part 1: Fitting a Multiple Regression

**Biological context**: plant ecologists studying *Arabidopsis thaliana* want to predict leaf chlorophyll content from three easily-measured variables — leaf area (cm²), nitrogen content (%), and days of growth.

```{r}
set.seed(42)
n <- 60

# Predictors
leaf_area <- rnorm(n, 8, 2.5)
nitrogen  <- rnorm(n, 3.2, 0.8)
days      <- runif(n, 14, 56)

# Response: chlorophyll (SPAD units), influenced by all three
chlorophyll <- 15 + 2.1 * leaf_area + 8.4 * nitrogen + 0.3 * days +
               rnorm(n, 0, 4)

arabidopsis <- data.frame(chlorophyll, leaf_area, nitrogen, days)

head(arabidopsis)
```

```{r}
# Fit multiple regression
model_full <- lm(chlorophyll ~ leaf_area + nitrogen + days,
                 data = arabidopsis)
summary(model_full)
```

::: callout-important
## Reading Multiple Regression Output

**Coefficients table:**

Each coefficient is a **partial regression coefficient** — the estimated change in y for a one-unit increase in that predictor, **holding all other predictors constant**.

For example: the coefficient for `nitrogen` estimates how much chlorophyll changes per 1% increase in N, *at fixed leaf area and days of growth*.

**Model fit:**

- `Multiple R-squared`: proportion of variance in chlorophyll explained by all predictors together
- `Adjusted R-squared`: penalises for the number of predictors (always ≤ R²). Use this to compare models with different numbers of predictors.
- `F-statistic`: tests H₀ that all slopes are zero simultaneously (overall model significance)
:::

------------------------------------------------------------------------

## Part 2: Comparing Models — Is Each Predictor Needed?

```{r}
# Is `days` a useful predictor, or is it noise?
model_reduced <- lm(chlorophyll ~ leaf_area + nitrogen, data = arabidopsis)

# Compare with ANOVA (nested model comparison)
anova(model_reduced, model_full)
```

```{r}
# AIC: lower = better model (balances fit vs complexity)
AIC(model_reduced, model_full)
```

::: callout-note
## AIC — Akaike Information Criterion

AIC penalises model complexity: AIC = 2k − 2 × log-likelihood, where k = number of parameters.

A difference in AIC (ΔAIC) of:
- < 2: models are similarly supported — prefer the simpler one
- 2–7: moderate evidence for the model with lower AIC
- > 10: strong evidence for the model with lower AIC

AIC does not give a p-value — it is a relative measure. You can only compare AIC between models fit to the **same dataset**.
:::

```{r}
# Stepwise model selection using AIC (automated)
library(MASS)
model_step <- stepAIC(model_full, direction = "both", trace = FALSE)
summary(model_step)
```

------------------------------------------------------------------------

## Part 3: Multicollinearity

If two predictors are highly correlated, the model cannot reliably separate their effects — a problem called **multicollinearity**. Coefficients become unstable and their standard errors inflate.

```{r}
# Check pairwise correlations among predictors
cor(arabidopsis[, c("leaf_area", "nitrogen", "days")])
```

```{r}
# Variance Inflation Factor (VIF) — formal multicollinearity measure
# install.packages("car")
library(car)
vif(model_full)
```

::: callout-warning
## Interpreting VIF

VIF measures how much the variance of each coefficient is inflated by its correlation with other predictors.

| VIF | Interpretation |
|-----|---------------|
| 1 | No multicollinearity |
| 1–5 | Moderate — generally acceptable |
| 5–10 | High — consider removing one correlated predictor |
| > 10 | Severe — model is unreliable |

If VIF is high: (1) remove one of the correlated predictors, (2) combine them into a composite score, or (3) use ridge regression.
:::

```{r}
# Demonstrate multicollinearity problem with highly correlated predictors
set.seed(7)
x1   <- rnorm(50)
x2   <- x1 + rnorm(50, 0, 0.2)   # x2 almost identical to x1
y_mc <- 2 * x1 + rnorm(50)

model_mc <- lm(y_mc ~ x1 + x2)
summary(model_mc)

cat("\nVIF:\n")
vif(model_mc)
```

------------------------------------------------------------------------

## Part 4: Regression Assumptions Revisited

```{r}
par(mfrow = c(2, 2))
plot(model_full)
par(mfrow = c(1, 1))
```

::: callout-tip
## What to Look For

**Residuals vs Fitted**: flat, randomly scattered around zero — no pattern. Curvature → missing non-linear terms. Funnel → heteroscedasticity.

**Normal Q-Q**: residuals on the line → normality satisfied.

**Scale-Location**: square root of |residuals| vs fitted — flat line confirms constant variance.

**Residuals vs Leverage**: points with high leverage (far from centre of x-space) and large residuals are **influential** — they can pull the regression line disproportionately. Cook's distance > 0.5 warrants investigation.
:::

------------------------------------------------------------------------

## Part 5: Categorical Predictors in Regression

```{r}
# Add a categorical predictor: growth condition (greenhouse vs field)
arabidopsis$condition <- factor(rep(c("Greenhouse", "Field"), each = 30))
arabidopsis$chlorophyll[arabidopsis$condition == "Field"] <-
  arabidopsis$chlorophyll[arabidopsis$condition == "Field"] - 5

# R automatically creates a dummy variable (0/1) for categorical predictors
model_with_condition <- lm(chlorophyll ~ leaf_area + nitrogen + days + condition,
                           data = arabidopsis)
summary(model_with_condition)
```

::: callout-note
## Categorical Predictors (Dummy Variables)

R converts a categorical factor with k levels into k − 1 dummy variables (0 or 1). The omitted level becomes the **reference category**, and its coefficient is absorbed into the intercept.

For a two-level factor (Greenhouse/Field), R creates one dummy: `conditionField = 1` if Field, 0 if Greenhouse. The coefficient for `conditionField` estimates the mean difference in chlorophyll between Field and Greenhouse plants, holding all other predictors constant.
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** A multiple regression predicting plant height from soil pH and temperature gives the coefficient for soil pH as 3.2 (p = 0.01). What does this mean?

**2.** A model with two predictors gives R² = 0.68 and adjusted R² = 0.65. Adding a third predictor changes these to R² = 0.69 and adjusted R² = 0.64. Should you keep the third predictor?

    a) Yes — R² increased    b) No — adjusted R² decreased, indicating the predictor is not useful enough to justify the added complexity    c) Yes — the model is always better with more variables    d) Cannot tell without a p-value

**3.** You fit a multiple regression and find VIF = 12 for two predictors. What does this mean and what should you do?

**4.** In a Residuals vs Leverage plot, one point has very high leverage but a small residual. Is this point necessarily a problem?

**5.** True or False: The intercept in a multiple regression is the predicted value of y when all predictor variables equal zero.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** Holding temperature constant, each one-unit increase in soil pH is associated with a 3.2 cm increase in plant height on average. The partial coefficient isolates the effect of pH while removing the influence of temperature. The p-value of 0.01 indicates this association is statistically significant.

**2.** b) No — adjusted R² decreased from 0.65 to 0.64, meaning the third predictor does not explain enough additional variance to justify the extra parameter. Adjusted R² penalises for model complexity; a decrease signals the predictor adds noise rather than signal. AIC comparison would confirm this.

**3.** VIF = 12 indicates **severe multicollinearity** — the two predictors are so highly correlated that the model cannot reliably estimate their separate effects. Standard errors of their coefficients are inflated, making them appear non-significant even if they truly matter. Solutions: (1) Remove one of the correlated predictors, (2) combine them into a single composite variable (e.g., by PCA), or (3) use regularised regression (ridge regression).

**4.** High leverage with a small residual means the point is at an unusual x-value but the model fits it well — it lies far from the centre of the predictor space but happens to fall close to the regression line. This is not necessarily a problem in itself. A point becomes truly **influential** (harmful) only when it has both high leverage AND a large residual, pulling the regression line toward itself. Cook's distance captures this combination.

**5.** True — the intercept (β₀) is the predicted value of y when every predictor variable in the model equals zero. Whether this is biologically meaningful depends on context: if x = 0 is impossible or outside the range of the data (e.g., body mass = 0), the intercept is a mathematical baseline rather than an interpretable quantity.
:::

------------------------------------------------------------------------

## Lab 25 Checklist

Before you leave, make sure you can:

- [ ] Fit `lm(y ~ x1 + x2 + x3)` and interpret partial regression coefficients
- [ ] Explain the difference between R² and adjusted R²
- [ ] Compare models with `AIC()` and interpret ΔAIC
- [ ] Use `cor()` to check for multicollinearity among predictors
- [ ] Interpret VIF values and identify when multicollinearity is a concern
- [ ] Run and interpret the four regression diagnostic plots
- [ ] Add a categorical predictor to `lm()` and interpret the dummy variable coefficient

::: callout-tip
## Bonus Challenge

Use the built-in R dataset `iris`. Fit three competing models predicting `Petal.Length`: (1) Sepal.Length alone, (2) Sepal.Length + Sepal.Width, (3) Sepal.Length + Sepal.Width + Species. Compare all three with AIC. Check VIF for model (2). Run diagnostic plots for model (3) and comment on whether the assumptions are met.
:::

------------------------------------------------------------------------

## Before Next Class (Wednesday)

- Wednesday (Lab 26): Measuring biodiversity in R — species richness, Shannon diversity, and Simpson's index using the vegan package
- Read **R for Data Science Ch. 24–25**

------------------------------------------------------------------------
