---
title: "Lab 26: Measuring Species Diversity in R — Part I"
subtitle: "BI 255 · Bethune-Cookman University · Fall 2026"
author: "Dr. Rosie Stanbrook-Buyer"
date: "October 28, 2026"
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    toc-title: "In This Lab"
    toc-depth: 3
    number-sections: false
    code-fold: false
    code-tools: true
    highlight-style: github
    smooth-scroll: true
    embed-resources: true
callout-appearance: default
execute:
  warning: false
  message: false
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

------------------------------------------------------------------------

## Quantifying Biodiversity

Biodiversity is more than just a count of species. A forest with 10 species, nine of which are rare and one of which makes up 95% of all individuals, is less diverse in a meaningful sense than a forest where 10 species are evenly distributed. **Diversity indices** capture both richness (how many species) and evenness (how equally abundant they are).

::: callout-note
## Learning Objectives
By the end of this lab you will be able to:

1. Distinguish between species richness, Shannon diversity, and Simpson's index
2. Install and use the `vegan` package for ecological data analysis
3. Compute diversity indices manually and with `vegan::diversity()`
4. Understand what evenness means and calculate Pielou's J
5. Construct a species abundance distribution plot
6. Compare diversity across sites and interpret differences biologically
:::

------------------------------------------------------------------------

## Part 1: Species Richness — The Simplest Measure

**Species richness (S)** is simply the number of species present in a sample. It is the most intuitive measure but ignores abundance distribution entirely.

```{r}
# Three Caribbean reef fish survey sites
# Columns = species, rows = sites
# Values = number of individuals observed

site_A <- c(Parrotfish = 45, Snapper = 42, Grouper = 40, Angelfish = 38,
            Tang = 35,  Wrasse = 30, Damselfish = 25, Triggerfish = 20)

site_B <- c(Parrotfish = 180, Snapper = 8, Grouper = 4, Angelfish = 3,
            Tang = 2, Wrasse = 2, Damselfish = 1, Triggerfish = 0)

site_C <- c(Parrotfish = 60, Snapper = 55, Grouper = 0, Angelfish = 0,
            Tang = 48, Wrasse = 0, Damselfish = 0, Triggerfish = 0)

cat("Species richness:\n")
cat("  Site A:", sum(site_A > 0), "species\n")
cat("  Site B:", sum(site_B > 0), "species\n")
cat("  Site C:", sum(site_C > 0), "species\n")
```

::: callout-warning
## The Problem with Richness Alone

Sites A and B both have 8 species — but intuitively, Site A is far more diverse because individuals are spread evenly across species. Site B is dominated by a single species (Parrotfish = 180 of the total).

Species richness misses this. Diversity indices that incorporate both richness and evenness give a more complete picture.
:::

------------------------------------------------------------------------

## Part 2: Shannon Diversity Index

The **Shannon index (H')** was originally derived from information theory. It measures the uncertainty in predicting the species identity of a randomly chosen individual. High H' means many species and/or even abundance distribution.

H' = −Σ(pᵢ × ln(pᵢ))

where pᵢ is the proportion of all individuals belonging to species i.

```{r}
# Manual calculation of Shannon's H'
shannon_manual <- function(counts) {
  counts <- counts[counts > 0]            # Remove absent species
  props  <- counts / sum(counts)          # Convert to proportions
  H      <- -sum(props * log(props))     # Shannon formula (natural log)
  return(H)
}

cat("Shannon H' (manual):\n")
cat("  Site A:", round(shannon_manual(site_A), 4), "\n")
cat("  Site B:", round(shannon_manual(site_B), 4), "\n")
cat("  Site C:", round(shannon_manual(site_C), 4), "\n")
```

```{r}
# Using the vegan package
# install.packages("vegan")
library(vegan)

community_matrix <- rbind(Site_A = site_A,
                           Site_B = site_B,
                           Site_C = site_C)

# Shannon diversity (vegan uses natural log by default)
diversity(community_matrix, index = "shannon")
```

::: callout-note
## Interpreting Shannon's H'

| H' value | Interpretation (rough guide) |
|----------|------------------------------|
| 0 | Only one species present |
| 1–2 | Low diversity |
| 2–3 | Moderate diversity |
| > 3 | High diversity |

H' has no absolute maximum — it depends on species richness. A community with S species and perfectly equal abundances achieves H'_max = ln(S). Always interpret H' relative to the maximum possible for that richness.
:::

------------------------------------------------------------------------

## Part 3: Simpson's Diversity Index

The **Simpson index** measures the probability that two individuals drawn randomly from the same sample belong to **different** species. It ranges from 0 (no diversity) to 1 (maximum diversity).

D = 1 − Σ(pᵢ²)

```{r}
# Simpson's index (1 - D in vegan notation)
diversity(community_matrix, index = "simpson")
```

```{r}
# Manual calculation for comparison
simpson_manual <- function(counts) {
  counts <- counts[counts > 0]
  props  <- counts / sum(counts)
  D      <- 1 - sum(props^2)
  return(D)
}

cat("Simpson 1-D (manual):\n")
cat("  Site A:", round(simpson_manual(site_A), 4), "\n")
cat("  Site B:", round(simpson_manual(site_B), 4), "\n")
cat("  Site C:", round(simpson_manual(site_C), 4), "\n")
```

::: callout-tip
## Shannon vs Simpson — Which to Use?

**Shannon H'** is more sensitive to rare species — adding even one rare species increases H' more than it increases Simpson's D. Use Shannon when rare species matter (e.g., conservation planning).

**Simpson D (1−D)** is dominated by abundant species and changes little when very rare species are added or removed. Use when you want a measure robust to sampling rare species.

Both are valid. Many ecologists report both, along with species richness.
:::

------------------------------------------------------------------------

## Part 4: Evenness — Pielou's J

**Pielou's J** rescales Shannon diversity relative to the theoretical maximum for that richness level:

J = H' / H'_max = H' / ln(S)

```{r}
# Calculate Pielou's J for all three sites
S <- specnumber(community_matrix)   # species richness per site
H <- diversity(community_matrix, index = "shannon")

J <- H / log(S)   # Pielou's evenness

cat("Species richness (S):\n"); print(S)
cat("\nShannon H':\n"); print(round(H, 3))
cat("\nPielou's J (evenness, 0–1):\n"); print(round(J, 3))
```

::: callout-note
## Interpreting Pielou's J

J ranges from 0 (all individuals in one species) to 1 (perfectly equal abundances across all species).

Site A: J ≈ 1.0 — nearly perfect evenness. Each species contributes similarly.

Site B: J ≈ 0.3 — low evenness despite equal richness. One species dominates.

Site C: J ≈ 1.0 — only 3 species but they are evenly abundant.

This illustrates why diversity = richness × evenness: Site A has both, Site C has evenness but not richness, Site B has richness but not evenness.
:::

------------------------------------------------------------------------

## Part 5: Species Abundance Distribution (SAD)

A **species abundance distribution** (rank-abundance curve) plots species in decreasing order of abundance. The shape of this curve reveals the overall evenness of a community.

```{r}
library(ggplot2)

# Prepare rank-abundance data
rad_data <- function(site_vec, site_name) {
  s  <- site_vec[site_vec > 0]
  s  <- sort(s, decreasing = TRUE)
  data.frame(
    Rank      = seq_along(s),
    Abundance = s,
    Species   = names(s),
    Site      = site_name
  )
}

rad_all <- rbind(rad_data(site_A, "Site A"),
                 rad_data(site_B, "Site B"),
                 rad_data(site_C, "Site C"))

ggplot(rad_all, aes(x = Rank, y = Abundance, colour = Site, group = Site)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  scale_colour_manual(values = c("Site A" = "#2C5F8A",
                                 "Site B" = "#C8102E",
                                 "Site C" = "#5B8DB8")) +
  labs(title    = "Rank-Abundance Curves for Three Caribbean Reef Fish Communities",
       subtitle  = "Flat curves = high evenness; steep curves = low evenness",
       x        = "Species Rank (most to least abundant)",
       y        = "Number of Individuals") +
  theme_classic(base_size = 13)
```

::: callout-note
## Reading a Rank-Abundance Curve

- **Flat, wide curve** (Site A): many species, similar abundances — high richness and high evenness
- **Steep, narrow curve** (Site B): one dominant species drops off sharply — low evenness
- **Short, flat curve** (Site C): few species but equal abundances — low richness, high evenness

The length of the curve on the x-axis shows richness. The steepness shows how unevenly distributed the individuals are.
:::

------------------------------------------------------------------------

## 3-Minute Knowledge Check

*Close your notes. Answer these on your own — you have 3 minutes. We'll go through the answers together after.*

::: callout-caution
## Knowledge Check Questions

**1.** Two forest plots each have 10 tree species. Plot 1 has 10 individuals of each species (100 total). Plot 2 has 91 individuals of one species and 1 of each of the other 9 species (100 total). Which plot has higher Shannon H' and why?

**2.** You calculate Simpson's 1−D for a grassland community and get 0.08. What does this tell you?

**3.** A site has H' = 2.1 and S = 15 species. What is Pielou's J, and is the community considered even?

**4.** On a rank-abundance curve, what does a long, nearly flat line indicate compared to a short, steep line?

**5.** True or False: It is possible for two communities to have the same species richness but very different Shannon diversity values.
:::

::: {.callout-caution collapse="true"}
## Answers (Only reveal if you have answered the above questions)

**1.** Plot 1 has higher Shannon H'. Shannon diversity is maximised when all species are equally abundant — Plot 1 has perfect evenness (J = 1.0). In Plot 2, the dominant species contributes −91/100 × ln(91/100) ≈ 0.09 to H', while the nine rare species each contribute only −1/100 × ln(1/100) ≈ 0.046 — resulting in a much lower total H'. Intuitively, you could almost predict the species identity of a random individual in Plot 2 (it's probably the dominant one), so uncertainty (= diversity) is low.

**2.** Simpson's 1−D = 0.08 is very low — close to 0 means very low diversity. The probability that two randomly drawn individuals come from different species is only 8%. This suggests the community is strongly dominated by one or a few species. This might indicate environmental stress, disturbance, or competitive dominance.

**3.** H'_max = ln(15) = 2.708. Pielou's J = 2.1 / 2.708 = **0.77**. A value of 0.77 is moderately high evenness — the community is reasonably well-distributed across species, reaching about 77% of the maximum possible diversity for 15 species. Not perfectly even but substantially more even than a dominated community.

**4.** A long, nearly flat rank-abundance curve indicates **high species richness with high evenness** — many species are present and their abundances are similar. A short, steep curve indicates **low richness and/or low evenness** — few species and/or a rapid decline from a dominant species to rare ones.

**5.** True — Shannon diversity depends on both richness AND evenness. Two communities with S = 10 species can have very different H' values if their abundance distributions differ. The community with even abundances will have H' near ln(10) ≈ 2.3; the community dominated by one species will have H' close to 0.
:::

------------------------------------------------------------------------

## Lab 26 Checklist

Before you leave, make sure you can:

- [ ] Explain the difference between species richness, Shannon H', Simpson D, and Pielou's J
- [ ] Install the `vegan` package and load it with `library(vegan)`
- [ ] Compute `diversity(matrix, index = "shannon")` and `diversity(matrix, index = "simpson")`
- [ ] Use `specnumber()` to get richness and calculate Pielou's J manually
- [ ] Plot a rank-abundance curve and interpret its shape
- [ ] Explain why a community with many rare species and one dominant species has low evenness

::: callout-tip
## Bonus Challenge

R's `vegan` package includes the `BCI` dataset — a classic census of tree species on 50 plots in Barro Colorado Island (Panama). Load it with `data(BCI)`. Calculate Shannon H', Simpson 1−D, and Pielou's J for each of the 50 plots. Plot the distribution of Shannon H' across plots as a histogram. Do the plots vary substantially in diversity? What ecological factors might explain this variation?
:::

------------------------------------------------------------------------

## Before Next Class (Monday, Week 12)

- Monday (Lab 27): Species Diversity Part II — rarefaction curves, diversity comparisons, and NMDS ordination
- Review vegan package documentation: `?diversity`, `?rarecurve`, `?metaMDS`

------------------------------------------------------------------------
